### 2024-10-18 paper.tex
Initial commit

\documentclass[10pt]{article}
\usepackage{sdss2020} % Uses Times Roman font (either newtx or times package)
\usepackage{url}
\usepackage{latexsym}
\usepackage{amsmath, amsthm, amsfonts}
\usepackage{algorithm, algorithmic}  
\usepackage{multirow,graphicx}

\title{Modeling Opening Weekend Revenue for U.S. Box Office Movies, 2005-2007}

\author{
  C. Logue \\
  Georgetown University \\
  Washington, DC \\
\\\And
  P. Mernagh \\
  Georgetown University \\
  Washington, DC \\
\\\And
  Z. Holden \\
  Georgetown University \\
  Washington, DC \\
\\}

\begin{document}
\maketitle

\section{Introduction}

\section{Data}

Data on approximately 500 movies that opened in U.S. theaters between 2005-2007 was aggregated from The Movie Database (TMDB). Available data points included the release date; a selection of genres that the movie could be classified as (e.g. Drama, Thriller, Action, Crime, Documentary, etc.); popularity, which is a proprietary TMDB metric based on various data points including votes, views, release date, etc.; vote average, the mean of user-submitted votes (on the scale of 0-10); vote count; budget; origin country or countries; production companies; production country or countries, runtime (in minutes); and what collection the movie belongs to, if any (e.g. the Harry Potter series). The dataset was augmented with opening weekend data from Box Office Mojo, a service of IMDB Pro, specifically opening revenue, opening revenue as a percent of total revenue, and the number of theaters the movie opened in.

The dataset is restricted to 2005-2007 to avoid adjusting for inflation. The time period is also chosen to predate the widespread adoption of streaming and its potential impacts on box office sales.

\section{Methods}

The list of possible predictors was narrowed to budget, number of production companies, runtime, opening weekend season, and whether the movie is part of a collection. 
The seasons are assigned based on the movie opening date falling into the following periods: Winter (January-February), Spring (March-April), Summer (May-August), Fall (September-October), and Holiday (November-December).

\section{Results}

\section{Discussion}

\bibliographystyle{sdss2020} % Please do not change the bibliography style
%\bibliography{fullbib}

\appendix

\section{Replication Data and Code}
\label{a:code}
Data and code are available at https://github.com/cons-code/MATH-5500
\end{document}

### 2024-10-19 paper.tex
Data discussion and summary stats

@@
 \section{Introduction}
-
+The project objective is to predict as accurately as possible opening weekend box office revenue for a set of movies, specifically from 2005-2007.
+Motivation for this project is a wealth of data points from both The Movie Database (TMDB) and Internet Movie Database (IMDb) around historical movies and TV shows: user-submitted ratings, applicable genres, budget, release date, actors, producer(s), production companies, country (or countries) of origin, etc.
+Further motivation is a prior understanding of likely useful predictors, such as the movie's budget, or the seasonal factors affecting a movie release.
+For example, the periods January-February, and to a lesser extent August-September, have been referred to as the ``dump months, '' i.e.
+effectively a dumping ground for low-potential movies that are not expected to make as much money.
+Various factors such as the seasonality of consumer spending (i.e binge before the holidays, pay down debt later), the timing of Golden Globes and Oscars awards, weather, the Super Bowl, and vacation months may all lead to seasonality in box office receipts.
+The years 2005-2007 were specifically chosen since they predate the rise of the streaming era and its potential impacts on box office sales.
+They are a short enough period that no inflation adjustment is made to any financial variables.
 \section{Data}
@@
 Data on approximately 500 movies that opened in U.S.
-theaters between 2005-2007 was aggregated from The Movie Database (TMDB).
-Available data points included the release date; a selection of genres that the movie could be classified as (e.g.
-Drama, Thriller, Action, Crime, Documentary, etc.); popularity, which is a proprietary TMDB metric based on various data points including votes, views, release date, etc.; vote average, the mean of user-submitted votes (on the scale of 0-10); vote count; budget; origin country or countries; production companies; production country or countries, runtime (in minutes); and what collection the movie belongs to, if any (e.g.
-the Harry Potter series).
-The dataset was augmented with opening weekend data from Box Office Mojo, a service of IMDB Pro, specifically opening revenue, opening revenue as a percent of total revenue, and the number of theaters the movie opened in.
+theaters between 2005-2007 was aggregated from TMDB and IMDb.
+Only English-language movies are included.
+Retrieved data points include the release date, the opening weekend revenue, budget, and the number of theaters the movie opened in.
+A selection of genres that each movie could be classified as (e.g.
+drama, thriller, action, crime, documentary, etc.) is included.
+Most movies are tagged with two to three genres, though a handful go as high as six.
+Popularity, a proprietary TMDB metric, is also included.
+It itself is based on various data points including votes, views, release date, etc..
+As a community-built database, TMDB provides the data point vote average, namely the mean of user-submitted ratings on the scale of 0-10, as well as the number of submitted votes.
+Both the origin and production country (or countries) are included.
+The production companies are provided (e.g big hitters such as Paramount, Warner Bros, MGM).
+Finally, the runtime, and what collection the movie belongs to, if any, such as the Harry Potter series or the Pirates of the Caribbean collection.
+The operating revenue and number of opening theaters were specifically pulled from Box Office Mojo, a service of IMDB Pro.
+The following predictors are chosen for initial modeling:
+\begin{enumerate}
+\item Budget (continuous)
+\item Number of opening theaters (continuous)
+\item Runtime (continuous)
+\item Number of production companies (continuous)
+\item From a series (binary, 118 total)
+\item Opening weekend season (categorical)
+\begin{enumerate}
+\item Winter: January-February (71 total)
+\item Spring: March-April (84 total)
+\item Summer: May-August (154 total)
+\item Fall: September-October (120 total)
+\item Holiday: November-December (82 total)
+\end{enumerate}
+\end{enumerate}
 
-The dataset is restricted to 2005-2007 to avoid adjusting for inflation. The time period is also chosen to predate the widespread adoption of streaming and its potential impacts on box office sales.
+Summary statistics are as follows:
+\begin{table}[H]
+\centering
+\caption{Summary Statistics (n = 511)}
+\begin{tabular}{|lrrr|} 
+\hline
+Variable & Mean & Min & Max \\ 
+\hline
+opening revenue (\$) & 14,743,982 & 9,653 & 151,116,516\\
+\hline
+budget (\$) & 40,365,469 & 0 & 300,000,000\\
+\hline
+\# of theaters & 2,134 & 1 & 4,362\\
+\hline
+runtime (min) & 107 & 14 & 187\\
+\hline
+\# of prod. cos. & 3.8 & 0 & 17\\
+\hline
+\end{tabular}  
+\end{table}
 
 \section{Methods}
-
-The list of possible predictors was narrowed to budget, number of production companies, runtime, opening weekend season, and whether the movie is part of a collection. 
-The seasons are assigned based on the movie opening date falling into the following periods: Winter (January-February), Spring (March-April), Summer (May-August), Fall (September-October), and Holiday (November-December).
 

### 2024-10-20 create_tmdb_model.py
add multivariate model functionality

@@ -1,0 +1,77 @@
+import argparse
+
+import numpy as np
+import pandas as pd
+import seaborn as sns
+
+from src.tmdb_frame import tmdb_frame
+from src.tmdb_model import tmdb_model
+from src.tmdb_linreg import tmdb_linreg
+
+
+#%% Main method & associated utilities
+
+# Parse input arguments.
+def parse_inputs():
+
+    # Initialize argument parser object.
+    parser = argparse.ArgumentParser(description='Tool to generate TMDB model from specified data file.')
+
+    # Add file parameters.
+    parser.add_argument('--fpath', type=str, default='./data', help='Relative filepath to input directory.')
+    parser.add_argument('--fname', type=str, default='first_cut_adj.csv', help='Name of input CSV file containing TMDB data.')
+    parser.add_argument('--n', type=int, default=500, help='Number of movies in TMDB data.')
+
+    # Parse input arguments.
+    args = parser.parse_args()
+
+    return args
+
+
+# Define main method.
+def main():
+
+    # Create GMTI model object from input arguments.
+    args = parse_inputs()
+
+    fpath = args.fpath
+    fname = args.fname
+
+    n = args.n
+
+    # Initialize TMDB model class.
+    mdl = tmdb_model()
+
+    # Generate cleaned TMDB dataframe.
+    df = tmdb_frame(fpath, fname, n=n)
+    cdf = df.get_clean_pd_df()
+
+    # Generate pair plot of continuous covariates.
+    sns.pairplot(cdf[['opening_revenue', 'budget', 'num_production_companies', 'runtime', 'popularity']])
+
+    # Generate OLS model.
+    Y = ['opening_revenue']
+    X = ['budget', 'popularity', 'runtime', 'is_series', 'is_fall', 'is_holiday']
+
+    cov_transform = {'opening_revenue': 'sqrt', 'runtime': 'sq'}
+    # cov_transform = None
+    
+    # cov_joint = [['is_series', 'is_holiday']]
+    cov_joint = None
+
+    _, res = mdl.get_ols_model(cdf, Y, X, cov_transform=cov_transform, cov_joint=cov_joint)
+
+    # Generate linear regression diagnostics for OLS model.
+    print(res.summary())
+
+    diag = tmdb_linreg(res)
+    vif, fig, ax = diag()
+
+
+#%% Main method execution
+
+if __name__ == '__main__':
+
+    main()
+
+    

### 2024-10-20 src/tmdb_frame.py

@@ -1,0 +1,135 @@
+import pandas as pd
+import numpy as np
+
+from utils.tmdb_utils import utils
+
+
+#%% TMDB frame class & associated utilities
+class tmdb_frame(utils):
+
+    # Initialize TMDB dataframe parameters.
+    def __init__(self, fpath: str, fname: str, n: int=500):
+
+        # Inherit parent utilities class.
+        super().__init__()
+        
+        # Initialize film parameter arrays.
+        self.title = np.zeros((n,), dtype=object)               # Movie title
+        self.title_id = np.zeros((n,), dtype=int)               # TMDB numerical identifier for movie title
+        self.imdb_id = np.zeros((n,), dtype=int)                # IMDB numerical identifier for movie title
+
+        self.runtime = np.zeros((n,), dtype=int)                # Total runtime (min)
+
+        self.is_series = np.zeros((n,), dtype=bool)             # Part of collection (T/F)
+        self.is_indie = np.zeros((n,), dtype=bool)              # Independent film (T/F)
+
+        self.num_studios = np.zeros((n,), dtype=int)            # Number of production companies
+        self.budget = np.zeros((n,), dtype=int)                 # Production budget ($USD)
+
+        # Initialize release date parameters.
+        self.is_spring = np.zeros((n,), dtype=bool)             # Released during the spring season (i.e., March-April) (T/F)
+        self.is_summer = np.zeros((n,), dtype=bool)             # Released during the summer season (i.e., May-August) (T/F)
+        self.is_fall = np.zeros((n,), dtype=bool)               # Released during the fall season (i.e., September-October) (T/F)
+        self.is_holiday = np.zeros((n,), dtype=bool)            # Released during the holiday season (i.e., November-December) (T/F)
+
+        # Initialize opening weekend parameters.
+        self.opening_revenue = np.zeros((n,), dtype=int)        # Opening weekend revenue ($USD)
+        self.num_opening_theaters = np.zeros((n,), dtype=int)   # Number of opening weekend theaters
+
+        # Initialize critical reception parameters.
+        self.popularity = np.zeros((n,), dtype=float)           # TDMB proprietary popularity rating
+
+        # Define TMDB items from dataframe.
+        df = self.parse_csv(fpath, fname)
+
+        for i in range(n):
+            
+            # Define film parameters.
+            self.title[i] = df.loc[i, 'release']
[…]

### 2024-10-20 src/tmdb_linreg.py

@@ -1,0 +1,340 @@
+import numpy as np
+import pandas as pd
+import seaborn as sns
+import matplotlib.pyplot as plt
+
+import statsmodels
+
+from statsmodels.tools.tools import maybe_unwrap_results
+from statsmodels.graphics.gofplots import ProbPlot
+from statsmodels.stats.outliers_influence import variance_inflation_factor
+
+from typing import Type
+
+style_talk = 'seaborn-talk'    #refer to plt.style.available
+
+
+#%% TMDB linear regression class & asociated utilities
+class tmdb_linreg():
+
+    """
+    Diagnostic plots to identify potential problems in a linear regression fit.
+    Mainly,
+        a. non-linearity of data
+        b. Correlation of error terms
+        c. non-constant variance
+        d. outliers
+        e. high-leverage points
+        f. collinearity
+
+    Authors:
+        Prajwal Kafle (p33ajkafle@gmail.com, where 3 = r)
+        Does not come with any sort of warranty.
+        Please test the code one your end before using.
+
+        Matt Spinelli (m3spinelli@gmail.com, where 3 = r)
+        (1) Fixed incorrect annotation of the top most extreme residuals in
+            the Residuals vs Fitted and, especially, the Normal Q-Q plots.
+        (2) Changed Residuals vs Leverage plot to match closer the y-axis
+            range shown in the equivalent plot in the R package ggfortify.
+        (3) Added horizontal line at y=0 in Residuals vs Leverage plot to
+            match the plots in R package ggfortify and base R.
+        (4) Added option for placing a vertical guideline on the Residuals
+            vs Leverage plot using the rule of thumb of h = 2p/n to denote
+            high leverage (high_leverage_threshold=True).
+        (5) Added two more ways to compute the Cook's Distance (D) threshold:
+            * 'baseR': D > 1 and D > 0.5 (default)
+            * 'convention': D > 4/n
+            * 'dof': D > 4 / (n - k - 1)
+        (6) Fixed class name to conform to Pascal casing convention
+        (7) Fixed Residuals vs Leverage legend to work with loc='best'
+    """
+
+    def __init__(self,
+                 results: Type[statsmodels.regression.linear_model.RegressionResultsWrapper]) -> None:
+        """
+        For a linear regression model, generates following diagnostic plots:
+
+        a. residual
+        b. qq
[…]

### 2024-10-20 utils/tmdb_utils.py

@@ -1,0 +1,231 @@
+import os
+import ast
+from typing import Tuple
+
+import numpy as np
+import pandas as pd
+
+
+#%% Built-in TMDB utilities class
+class utils():
+
+    # Initialize utility parameters.
+    def __init__(self):
+
+        self.major_studio_id = np.array([33,                                # Universal Pictures
+                                         3436,                              # Universal Studios Home Entertainment Family Productions
+                                         5556,                              # Universal Animation Studios
+                                         47046,                             # Universal Pictures do Brazil
+                                         4,                                 # Paramount Pictures
+                                         838,                               # Paramount Vantage
+                                         21834,                             # Paramount Pictures Digital Entertainment
+                                         220287,                            # Paramount Classics
+                                         174,                               # Warner Bros. Pictures
+                                         2785,                              # Warner Bros. Animation
+                                         17466,                             # Warner Bros. Entertainment Germany
+                                         12,                                # New Line Cinema
+                                         2,                                 # Walt Disney Pictures
+                                         6125,                              # Walt Disney Animation Studios
+                                         178202,                            # Walt Disney Pictures China
+                                         25,                                # 20th Century Fox
+                                         3,                                 # Pixar
+                                         34,                                # Sony Pictures
+                                         58,                                # Sony Pictures Classics
+                                         2251,                              # Sony Pictures Animation
+                                         5388,                              # Sony Pictures Home Entertainment
+                                         5,                                 # Columbia Pictures
+                                         559],                              # TriStar Pictures
+                                         dtype=np.int32)
+
+
+    # Convert string to Boolean.
+    def str_to_bool(self, string: str) -> bool:
+
+        if (string.lower() == 'false'):
+
[…]

### 2024-10-23 paper.tex
Data and methods update

@@
 \usepackage{multirow,graphicx}
+\usepackage[flushleft]{threeparttable}
+\usepackage[T1]{fontenc}
+\graphicspath{ {./images/} }
 
@@
 \item Budget (continuous)
-\item Number of opening theaters (continuous)
 \item Runtime (continuous)
@@
 \hline
-\# of theaters & 2,134 & 1 & 4,362\\
-\hline
 runtime (min) & 107 & 14 & 187\\
@@
 
+The variable popularity is not considered, since it is modeled by TMDB and likely influenced by post-opening weekend factors.
+The number of theaters the movie opened in is likely highly correlated with opening revenue.
+However, the project prefers variables that are intrinsic characteristics of a movie, thus number of theaters is also not considered further.
+
+A secondary set of variables, intended for follow-up modeling, is also aggregated or created via transformation.
+These include data points around the production room, writers room, sound room, and crew.
+Additionally, data around the credits and median rating is provided for producers, writers, and leads.
+A binary variable for whether the movie is based on a novel is included.
+An indie binary variable is created based on whether the movie was produced by a major production company or by a smaller (indie) company.
+Finally, indicator variables for the following genres are created: adventure, fantasy, animation, drama, horror, action, comedy, history, western, thriller, crime, documentary, science fiction, mystery, music, romance, family, and war.
+
 \section{Methods}
+First, the relationships between the initial selected predictors are visually examined.
+From pair plots of the continuous predictors (Figure 1), budget and opening revenue seem positively, linearly related.
+There is no apparent relationship between opening revenue and the number of production companies or the runtime of the movie.
+
+For the non-continous predictors, the Spring, Summer, and Holiday seasons are associated with higher opening revenue on average (Figure 2), though this is skewed, especially by releases grossing over \$50M.
+Movies that are part of a series also are associated with higher opening revenue on average (Figure 3).
+
+Armed with this knowledge, various models are fit to the data.
+Initially, a model incorporating all variables produces $R^2_a$ of over 60\%, but demonstrates clear heteroskedasticity of residuals.
+A square-root transformation noticeably stabilizes the variance.
+After this transformation, the initial best model candidate is:
+
+\begin{table}[H]
+  \begin{threeparttable}
+    \caption{Initial Model Specification}
+       \begin{tabular}{|l l l l l|} 
+	 \hline
+	 & Coef & 95\% CI & t-stat & p-value \\ [0.5ex] 
+	 \hline
+	intercept & 2006.7 & [1810.3, 2203.1] & 20.075 & <0.0001 \\ 
+	\hline
+	budget & 2.96e-05 & [2.67e-05, 3.25e-05] & 20.020 & <0.0001\\
+	 \hline
+	is\_series & 1379.9 & [1088.4, 1671.5] & 9.298 & <0.0001\\
+	\hline
+	is\_fall & -795.2 & [-1085.1, -505.4] & -5.390 & <0.0001\\
+	\hline
+	is\_holiday & -729.1 & [-1063.7, -394.5] & -4.281 & <0.0001\\[1ex]
+	\hline
+      \end{tabular}
+    \begin{tablenotes}
+      \small
+      \item $R^2_a$ = 0.58
+    \end{tablenotes}  
+  \end{threeparttable}    
+\end{table}
+
+Backwards elimination:
 
@@
 
-\bibliographystyle{sdss2020} % Please do not change the bibliography style
+\bibliographystyle{sdss2020} 
 %\bibliography{fullbib}
@@
 
+\section{Visualizations}
+\begin{figure}[H]
+	\begin{center}
+		\centerline{\includegraphics[width=\columnwidth]{pairs_plot}}
+		\caption{Pairs plot, initially selected continuous variables.}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+		\centerline{\includegraphics[scale=0.5]{season_revenue}}
+		\caption{Distribution of opening revenue by season}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+		\centerline{\includegraphics[scale=0.6]{is_series}}
+		\caption{Distribution of opening revenue, series vs. non-series}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+		\centerline{\includegraphics[width=\columnwidth]{cross_val}}
+		\caption{Cross-validation prediction results.}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+		\centerline{\includegraphics[width=\columnwidth]{partial_regression}}
+		\caption{Partial Regression Plots}
+	\end{center}
+\end{figure}
+
 \section{Replication Data and Code}
 \label{a:code}
-Data and code are available at https://github.com/cons-code/MATH-5500
+The repo \mbox{https://github.com/cons-code/MATH-5500} contains replication data and code.
 \end{document}

### 2024-10-23 paper.tex
Initial model and influential points

@@
 \usepackage[T1]{fontenc}
+\usepackage{booktabs}
+\usepackage{siunitx} 
 \graphicspath{ {./images/} }
 
-\title{Modeling Opening Weekend Revenue for U.S. Box Office Movies, 2005-2007}
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+
+\title{Modeling Opening Weekend Box Office Revenue for U.S. Movies, 2005-2007}
 
@@
 Various factors such as the seasonality of consumer spending (i.e binge before the holidays, pay down debt later), the timing of Golden Globes and Oscars awards, weather, the Super Bowl, and vacation months may all lead to seasonality in box office receipts.
-The years 2005-2007 were specifically chosen since they predate the rise of the streaming era and its potential impacts on box office sales.
+The years 2005-2007 are specifically chosen since they predate the rise of the streaming era and its potential impacts on box office sales.
 They are a short enough period that no inflation adjustment is made to any financial variables.
@@
 \caption{Summary Statistics (n = 511)}
-\begin{tabular}{|lrrr|} 
-\hline
+\begin{tabular}{lrrr} 
+\toprule
 Variable & Mean & Min & Max \\ 
-\hline
+\midrule
 opening revenue (\$) & 14,743,982 & 9,653 & 151,116,516\\
-\hline
 budget (\$) & 40,365,469 & 0 & 300,000,000\\
-\hline
 runtime (min) & 107 & 14 & 187\\
-\hline
 \# of prod. cos. & 3.8 & 0 & 17\\
-\hline
+\bottomrule
 \end{tabular}  
@@
 A secondary set of variables, intended for follow-up modeling, is also aggregated or created via transformation.
-These include data points around the production room, writers room, sound room, and crew.
+These include data points around the number of people working in the production room, writers room, and sound room, as well as number of crew.
 Additionally, data around the credits and median rating is provided for producers, writers, and leads.
@@
 \section{Methods}
-First, the relationships between the initial selected predictors are visually examined.
+The relationships between the initial selected predictors and opening revenue are visually examined.
 From pair plots of the continuous predictors (Figure 1), budget and opening revenue seem positively, linearly related.
@@
 
-For the non-continous predictors, the Spring, Summer, and Holiday seasons are associated with higher opening revenue on average (Figure 2), though this is skewed, especially by releases grossing over \$50M.
+For the non-continous predictors, the spring, summer, and holiday seasons are associated with higher opening revenue on average (Figure 2), though this is skewed, especially by releases grossing over \$50M.
 Movies that are part of a series also are associated with higher opening revenue on average (Figure 3).
@@
 Armed with this knowledge, various models are fit to the data.
-Initially, a model incorporating all variables produces $R^2_a$ of over 60\%, but demonstrates clear heteroskedasticity of residuals.
+Initially, a model incorporating all variables produces $R^2_a$ of over 0.6, but demonstrates clear heteroskedasticity of residuals.
 A square-root transformation noticeably stabilizes the variance.
-After this transformation, the initial best model candidate is:
+Runtime and number of production companies are dropped from consideration.
+This establishes a baseline model, which we will then attempt to improve with variables from our secondary set.
+Both DFFITS (Table 3, Figure 5) and DFBETAS (Tables 4-5) analyses are performed to identify potentially influential observations.
+
+\section{Results}
+
+The initial model specification is:
 
@@
     \caption{Initial Model Specification}
-       \begin{tabular}{|l l l l l|} 
-	 \hline
-	 & Coef & 95\% CI & t-stat & p-value \\ [0.5ex] 
-	 \hline
-	intercept & 2006.7 & [1810.3, 2203.1] & 20.075 & <0.0001 \\ 
-	\hline
-	budget & 2.96e-05 & [2.67e-05, 3.25e-05] & 20.020 & <0.0001\\
-	 \hline
-	is\_series & 1379.9 & [1088.4, 1671.5] & 9.298 & <0.0001\\
-	\hline
-	is\_fall & -795.2 & [-1085.1, -505.4] & -5.390 & <0.0001\\
-	\hline
-	is\_holiday & -729.1 & [-1063.7, -394.5] & -4.281 & <0.0001\\[1ex]
-	\hline
+       \begin{tabular}{l r r r r} 
+	 \toprule
+	 & Coef. & 95\% CI & t-stat & p-value \\ [0.5ex] 
+	 \midrule
+	intercept & 2162.6 & [1832.8, 2492.4] & 12.882 & <0.0001 \\ 
+	budget & 3e-05 & [2.7e-05, 3.3e-05] & 19.856 & <0.0001\\
+	is\_series & 1374.8 & [1083.0, 1666.6] & 9.255 & <0.0001\\
+	is\_spring & -152.8 & [-583.4, 277.8] & -0.697 & 0.486\\
+	is\_summer & -257.5 & [-646.8, 131.8] & -1.300 & 0.194\\
+	is\_fall & -960.5 & [-1360.5, -560.5] & -4.718 & <0.0001\\
+	is\_holiday & -904.0 & [-1344.7, -463.4] & -4.030 & <0.0001\\[1ex]
+	\bottomrule
       \end{tabular}
@@
 
-Backwards elimination:
-
-\section{Results}
-
 \section{Discussion}
@@
 %\bibliography{fullbib}
-
 \appendix
@@
 	\begin{center}
-		\centerline{\includegraphics[width=\columnwidth]{cross_val}}
-		\caption{Cross-validation prediction results.}
+		\centerline{\includegraphics[scale=0.4]{initial_diagnostics}}
+		\caption{Residual diagnostics, initial model}
+	\end{center}
+\end{figure}
+
+\begin{figure}[H]
+	\begin{center}
+		\centerline{\includegraphics[width=\columnwidth]{influence_plot}}
+		\caption{Influence plot (size based on DFFITS)}
 	\end{center}
@@
 
+\section{Analyses}
+
+\begin{table}[H]
+\caption{DFFITS results}
+\begin{tabular}{llr}
+\toprule
+obs & release & dffits \\
+\midrule
+281 & Superman Returns & -0.42 \\
+176 & Basic Instinct 2 & -0.38 \\
+509 & Wild Hogs & 0.38 \\
+102 & Son of the Mask & -0.37 \\
+47 & Harry Potter and the Goblet of Fire & 0.37 \\
+380 & Evan Almighty & -0.36 \\
+438 & Pirates of the Caribbean: At World's End & -0.35 \\
+103 & Star Wars: Episode III - Revenge of the Sith & 0.34 \\
+236 & Letters from Iwo Jima & -0.34 \\
+452 & Shrek the Third & 0.33 \\
+257 & Pirates of the Caribbean: Dead Man's Chest & 0.33 \\
+339 & 300 & 0.31 \\
+431 & Norbit & 0.31 \\
+493 & The Simpsons Movie & 0.31 \\
+97 & Saw II & 0.30 \\
+266 & Saw III & 0.29 \\
+168 & An Inconvenient Truth & -0.28 \\
+449 & Saw IV & 0.28 \\
+5 & Alone in the Dark & -0.27 \\
+74 & Memoirs of a Geisha & -0.27 \\
+458 & Superbad & 0.27 \\
+457 & Sunshine & -0.26 \\
+116 & The Exorcism of Emily Rose & 0.26 \\
+395 & Happily N'Ever After & -0.26 \\
+224 & Ice Age: The Meltdown & 0.25 \\
+494 & The Ultimate Gift & -0.25 \\
+228 & Jackass Number Two & 0.25 \\
+465 & The Bourne Ultimatum & 0.25 \\
+198 & Dreamgirls & -0.24 \\
+2 & A Sound of Thunder & -0.24 \\
+346 & Alvin and the Chipmunks & 0.23 \\
+454 & Spider-Man 3 & 0.23 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\caption{DFBETAs, budget (top 10 observations)}
+\begin{tabular}[t]{l L{0.5\linewidth} rr}
+\toprule
+obs & release & budget & dfbeta \\
+\midrule
+281 & Superman Returns & 223000000 & -0.35 \\
+438 & Pirates of the Caribbean: At World's End & 300000000 & -0.32 \\
+257 & Pirates of the Caribbean: Dead Man's Chest & 200000000 & 0.26 \\
+380 & Evan Almighty & 175000000 & -0.26 \\
+452 & Shrek the Third & 160000000 & 0.22 \\
+454 & Spider-Man 3 & 258000000 & 0.21 \\
+258 & Poseidon & 160000000 & -0.19 \\
+47 & Harry Potter and the Goblet of Fire & 150000000 & 0.19 \\
+105 & Stealth & 135000000 & -0.17 \\
+472 & The Golden Compass & 180000000 & -0.17 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\caption{DFBETAs, is\_series (top 10 observations)}
+\begin{tabular}[t]{l L{0.5\linewidth} rr}
+\toprule
+obs & release & is\_series & dfbeta \\
+\midrule
+236 & Letters from Iwo Jima & 1 & -0.22 \\
+97 & Saw II & 1 & 0.21 \\
+266 & Saw III & 1 & 0.21 \\
+449 & Saw IV & 1 & 0.20 \\
+168 & An Inconvenient Truth & 1 & -0.20 \\
+176 & Basic Instinct 2 & 1 & -0.20 \\
+228 & Jackass Number Two & 1 & 0.18 \\
+103 & Star Wars: Episode III - Revenge of the Sith & 1 & 0.17 \\
+339 & 300 & 1 & 0.17 \\
+102 & Son of the Mask & 1 & -0.17 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \section{Replication Data and Code}

### 2024-10-24 create_tmdb_model.py
add third cut features and covariate interactions

@@ -19,7 +19,7 @@ def parse_inputs():
 
     # Add file parameters.
     parser.add_argument('--fpath', type=str, default='./data', help='Relative filepath to input directory.')
-    parser.add_argument('--fname', type=str, default='first_cut_adj.csv', help='Name of input CSV file containing TMDB data.')
+    parser.add_argument('--fname', type=str, default='final_cut.csv', help='Name of input CSV file containing TMDB data.')
     parser.add_argument('--n', type=int, default=500, help='Number of movies in TMDB data.')
 
     # Parse input arguments.
@@ -47,17 +47,22 @@ def main():
     cdf = df.get_clean_pd_df()
 
     # Generate pair plot of continuous covariates.
-    sns.pairplot(cdf[['opening_revenue', 'budget', 'num_production_companies', 'runtime', 'popularity']])
+    sns.pairplot(cdf[['opening_revenue', 'budget', 'num_production_companies', 'runtime', 'month', 'pct_indie']])
 
     # Generate OLS model.
     Y = ['opening_revenue']
-    X = ['budget', 'popularity', 'runtime', 'is_series', 'is_fall', 'is_holiday']
+    X = ['budget', 'runtime', 'writers_room_size', 'num_production_companies', \
+         'pct_indie', 'is_indie', \
+         'cast_xp_sum', 'director_xp', 'director_rating_median', \
+         'prev_film_release', 'prev_film_rating', \
+         'is_feb', 'is_mar', 'is_oct', 'is_dec', \
+         'is_drama', 'is_horror', 'is_comedy', 'is_western', 'is_crime', 'is_documentary', 'is_sci_fi', 'is_music', 'is_family']
 
-    cov_transform = {'opening_revenue': 'sqrt', 'runtime': 'sq'}
+    cov_transform = {'opening_revenue': 'sqrt'}
     # cov_transform = None
     
-    # cov_joint = [['is_series', 'is_holiday']]
-    cov_joint = None
+    cov_joint = [['prev_film_release', 'prev_film_rating'], ['is_horror', 'is_oct'], ['cast_xp_sum', 'director_xp']]
+    # cov_joint = None
 
     _, res = mdl.get_ols_model(cdf, Y, X, cov_transform=cov_transform, cov_joint=cov_joint)
 

### 2024-10-24 src/tmdb_frame.py

@@ -13,31 +13,70 @@ class tmdb_frame(utils):
         # Inherit parent utilities class.
         super().__init__()
         
-        # Initialize film parameter arrays.
-        self.title = np.zeros((n,), dtype=object)               # Movie title
-        self.title_id = np.zeros((n,), dtype=int)               # TMDB numerical identifier for movie title
-        self.imdb_id = np.zeros((n,), dtype=int)                # IMDB numerical identifier for movie title
+        # Initialize film parameters.
+        self.title = np.zeros((n,), dtype=object)                   # Movie title
+        self.title_id = np.zeros((n,), dtype=int)                   # TMDB numerical identifier for movie title
+        self.imdb_id = np.zeros((n,), dtype=int)                    # IMDB numerical identifier for movie title
 
-        self.runtime = np.zeros((n,), dtype=int)                # Total runtime (min)
+        self.runtime = np.zeros((n,), dtype=int)                    # Total runtime (min)
 
-        self.is_series = np.zeros((n,), dtype=bool)             # Part of collection (T/F)
-        self.is_indie = np.zeros((n,), dtype=bool)              # Independent film (T/F)
+        self.is_series = np.zeros((n,), dtype=bool)                 # Part of collection (T/F)
+        self.prev_film_release = np.zeros((n,), dtype=bool)         # Not first movie in collection (T/F)
+        self.prev_film_rating = np.zeros((n,), dtype=float)         # Rating of previous movie in collection (T/F)
+        self.is_indie = np.zeros((n,), dtype=bool)                  # Independent film (T/F)
+        self.is_based_on_novel = np.zeros((n,), dtype=bool)         # Based on novel (T/F)
 
-        self.num_studios = np.zeros((n,), dtype=int)            # Number of production companies
-        self.budget = np.zeros((n,), dtype=int)                 # Production budget ($USD)
+        self.num_studios = np.zeros((n,), dtype=int)                # Number of production companies
+        self.pct_indie = np.zeros((n,), dtype=float)                # Percentage of independent production companies
+        self.budget = np.zeros((n,), dtype=int)                     # Production budget ($USD)
+
+        # Initialize production parameters.
+        self.writer_xp_median = np.zeros((n,), dtype=float)         # Writer credits median
[…]

### 2024-10-26 03_Methodology_10-25/src/tmdb_frame.py
hey

@@ -1,0 +1,291 @@
+import pandas as pd
+import numpy as np
+
+from utils.tmdb_utils import utils
+
+
+#%% TMDB frame class & associated utilities
+class tmdb_frame(utils):
+
+    # Initialize TMDB dataframe parameters.
+    def __init__(self, fpath: str, fname: str, n: int=500):
+
+        # Inherit parent utilities class.
+        super().__init__()
+        
+        # Initialize film parameters.
+        self.title = np.zeros((n,), dtype=object)                   # Movie title
+        self.title_id = np.zeros((n,), dtype=int)                   # TMDB numerical identifier for movie title
+        #self.imdb_id = np.zeros((n,), dtype=int)                    # IMDB numerical identifier for movie title
+
+        self.runtime = np.zeros((n,), dtype=int)                    # Total runtime (min)
+
+        self.is_series = np.zeros((n,), dtype=bool)                 # Part of collection (T/F)
+        self.prev_film_release = np.zeros((n,), dtype=bool)         # Not first movie in collection (T/F)
+        self.prev_film_rating = np.zeros((n,), dtype=float)         # Rating of previous movie in collection (T/F)
+        self.is_indie = np.zeros((n,), dtype=bool)                  # Independent film (T/F)
+        self.is_based_on_novel = np.zeros((n,), dtype=bool)         # Based on novel (T/F)
+
+        self.num_studios = np.zeros((n,), dtype=int)                # Number of production companies
+        self.pct_indie = np.zeros((n,), dtype=float)                # Percentage of independent production companies
+        self.budget = np.zeros((n,), dtype=int)                     # Production budget ($USD)
+
+        # Initialize production parameters.
+        self.writer_xp_median = np.zeros((n,), dtype=float)         # Writer credits median
+        self.writer_xp_sum = np.zeros((n,), dtype=int)              # Writer credits sum
+
+        self.writer_rating_median = np.zeros((n,), dtype=float)     # Writer rating median
+        self.writer_rating_max = np.zeros((n,), dtype=int)          # Writer rating maximum
+
+        self.director_xp = np.zeros((n,), dtype=int)                # Director credits
+
+        self.director_rating_median = np.zeros((n,), dtype=float)   # Director rating median
[…]

### 2024-10-26 03_Methodology_10-25/src/tmdb_frame.py
Setting up best model with interaction engineering at top, help with iteration a little more

@@ -1,291 +1,0 @@
-import pandas as pd
-import numpy as np
-
-from utils.tmdb_utils import utils
-
-
-#%% TMDB frame class & associated utilities
-class tmdb_frame(utils):
-
-    # Initialize TMDB dataframe parameters.
-    def __init__(self, fpath: str, fname: str, n: int=500):
-
-        # Inherit parent utilities class.
-        super().__init__()
-        
-        # Initialize film parameters.
-        self.title = np.zeros((n,), dtype=object)                   # Movie title
-        self.title_id = np.zeros((n,), dtype=int)                   # TMDB numerical identifier for movie title
-        #self.imdb_id = np.zeros((n,), dtype=int)                    # IMDB numerical identifier for movie title
-
-        self.runtime = np.zeros((n,), dtype=int)                    # Total runtime (min)
-
-        self.is_series = np.zeros((n,), dtype=bool)                 # Part of collection (T/F)
-        self.prev_film_release = np.zeros((n,), dtype=bool)         # Not first movie in collection (T/F)
-        self.prev_film_rating = np.zeros((n,), dtype=float)         # Rating of previous movie in collection (T/F)
-        self.is_indie = np.zeros((n,), dtype=bool)                  # Independent film (T/F)
-        self.is_based_on_novel = np.zeros((n,), dtype=bool)         # Based on novel (T/F)
-
-        self.num_studios = np.zeros((n,), dtype=int)                # Number of production companies
-        self.pct_indie = np.zeros((n,), dtype=float)                # Percentage of independent production companies
-        self.budget = np.zeros((n,), dtype=int)                     # Production budget ($USD)
-
-        # Initialize production parameters.
-        self.writer_xp_median = np.zeros((n,), dtype=float)         # Writer credits median
-        self.writer_xp_sum = np.zeros((n,), dtype=int)              # Writer credits sum
-
-        self.writer_rating_median = np.zeros((n,), dtype=float)     # Writer rating median
-        self.writer_rating_max = np.zeros((n,), dtype=int)          # Writer rating maximum
-
-        self.director_xp = np.zeros((n,), dtype=int)                # Director credits
-
-        self.director_rating_median = np.zeros((n,), dtype=float)   # Director rating median
[…]

### 2024-10-26 03_Methodology_10-25/src/tmdb_linreg.py

@@ -1,340 +1,0 @@
-import numpy as np
-import pandas as pd
-import seaborn as sns
-import matplotlib.pyplot as plt
-
-import statsmodels
-
-from statsmodels.tools.tools import maybe_unwrap_results
-from statsmodels.graphics.gofplots import ProbPlot
-from statsmodels.stats.outliers_influence import variance_inflation_factor
-
-from typing import Type
-
-style_talk = 'seaborn-talk'    #refer to plt.style.available
-
-
-#%% TMDB linear regression class & asociated utilities
-class tmdb_linreg():
-
-    """
-    Diagnostic plots to identify potential problems in a linear regression fit.
-    Mainly,
-        a. non-linearity of data
-        b. Correlation of error terms
-        c. non-constant variance
-        d. outliers
-        e. high-leverage points
-        f. collinearity
-
-    Authors:
-        Prajwal Kafle (p33ajkafle@gmail.com, where 3 = r)
-        Does not come with any sort of warranty.
-        Please test the code one your end before using.
-
-        Matt Spinelli (m3spinelli@gmail.com, where 3 = r)
-        (1) Fixed incorrect annotation of the top most extreme residuals in
-            the Residuals vs Fitted and, especially, the Normal Q-Q plots.
-        (2) Changed Residuals vs Leverage plot to match closer the y-axis
-            range shown in the equivalent plot in the R package ggfortify.
-        (3) Added horizontal line at y=0 in Residuals vs Leverage plot to
-            match the plots in R package ggfortify and base R.
-        (4) Added option for placing a vertical guideline on the Residuals
-            vs Leverage plot using the rule of thumb of h = 2p/n to denote
-            high leverage (high_leverage_threshold=True).
-        (5) Added two more ways to compute the Cook's Distance (D) threshold:
-            * 'baseR': D > 1 and D > 0.5 (default)
-            * 'convention': D > 4/n
-            * 'dof': D > 4 / (n - k - 1)
-        (6) Fixed class name to conform to Pascal casing convention
-        (7) Fixed Residuals vs Leverage legend to work with loc='best'
-    """
-
-    def __init__(self,
-                 results: Type[statsmodels.regression.linear_model.RegressionResultsWrapper]) -> None:
-        """
-        For a linear regression model, generates following diagnostic plots:
-
-        a. residual
-        b. qq
[…]

### 2024-10-26 03_Methodology_10-25/src/tmdb_model.py

@@ -1,140 +1,0 @@
-from typing import Tuple
-
-import numpy as np
-import pandas as pd
-
-from statsmodels.base.model import LikelihoodModel
-from statsmodels.base.model import Results
-
-from statsmodels.formula.api import ols
-from statsmodels.regression.linear_model import OLS
-from statsmodels.tools.tools import add_constant
-
-from utils.tmdb_utils import utils
-
-
-# TMDB model class & associated utilities
-class tmdb_model(utils):
-
-    # Initialize TMDB model parameters.
-    def __init__(self):
-
-        # Inherit parent utilities class.
-        super().__init__()
-    
-
-    # Generate OLS model to fit Y from predictor(s) X.
-    def get_ols_model(self, df: pd.DataFrame, Y: list, X: list, \
-                      cov_transform: dict=None, cov_joint: list=None) -> Tuple[LikelihoodModel, Results]:
-
-        # Define independent and dependent covariates.
-        Y = df[Y]
-
-        X = df[X]
-        X = add_constant(X)     # Add column of ones for intercept
-
-        # Calculate covariate transformations.
-        if bool(cov_transform):
-
-            for (key, val) in cov_transform.items():
-                
-                try:
-
-                    c = np.array(X[key])
-                    ct = self.get_cov_transform(c, shape=val)
-
-                    X[str(val + '_' + key)] = ct
-
-                except(KeyError):
-
-                    c = np.array(Y[key])
-                    ct = self.get_cov_transform(c, shape=val)
-
-                    Y[str(val + '_' + key)] = ct
-
-                    Y.drop(key, axis=1, inplace=True)
-
-        # Calculate joint covariates.
-        if bool(cov_joint):
-
-            for val in cov_joint:
-
-                n = len(val)
-
-                if (n == 2):
-
-                    ct = self.get_cov_joint(np.array(X[val[0]]), np.array(X[val[1]]))
-                    X[str(val[0] + '*' + val[1])] = ct
-
-                else:
-
-                    ct = self.get_cov_joint(np.array(X[val[0]]), np.array(X[val[1]]), np.array(X[val[2]]))
-                    X[str(val[0] + '*' + val[1] + '*' + val[2])] = ct
-
-        # Generate and fit OLS model.
-        model = OLS(Y, X)
-        res = model.fit()
-
-        return (model, res)
-    
-
-    # Generate OLS model to fit Y from predictor(s) X using R interface.
[…]

### 2024-10-26 03_Methodology_10-25/utils/tmdb_utils.py

@@ -1,231 +1,0 @@
-import os
-import ast
-from typing import Tuple
-
-import numpy as np
-import pandas as pd
-
-
-#%% Built-in TMDB utilities class
-class utils():
-
-    # Initialize utility parameters.
-    def __init__(self):
-
-        self.major_studio_id = np.array([33,                                # Universal Pictures
-                                         3436,                              # Universal Studios Home Entertainment Family Productions
-                                         5556,                              # Universal Animation Studios
-                                         47046,                             # Universal Pictures do Brazil
-                                         4,                                 # Paramount Pictures
-                                         838,                               # Paramount Vantage
-                                         21834,                             # Paramount Pictures Digital Entertainment
-                                         220287,                            # Paramount Classics
-                                         174,                               # Warner Bros. Pictures
-                                         2785,                              # Warner Bros. Animation
-                                         17466,                             # Warner Bros. Entertainment Germany
-                                         12,                                # New Line Cinema
-                                         2,                                 # Walt Disney Pictures
-                                         6125,                              # Walt Disney Animation Studios
-                                         178202,                            # Walt Disney Pictures China
-                                         25,                                # 20th Century Fox
-                                         3,                                 # Pixar
-                                         34,                                # Sony Pictures
-                                         58,                                # Sony Pictures Classics
-                                         2251,                              # Sony Pictures Animation
-                                         5388,                              # Sony Pictures Home Entertainment
-                                         5,                                 # Columbia Pictures
-                                         559],                              # TriStar Pictures
-                                         dtype=np.int32)
-
-
-    # Convert string to Boolean.
-    def str_to_bool(self, string: str) -> bool:
-
-        if (string.lower() == 'false'):
-
[…]

### 2024-11-02 paper.tex
Minor revision

--- 2024-10-19 paper.tex
@@
 \usepackage{multirow,graphicx}
+\usepackage[flushleft]{threeparttable}
+\usepackage[T1]{fontenc}
+\usepackage{booktabs}
+\usepackage{siunitx} 
+\graphicspath{ {./Images/} }
 
-\title{Modeling Opening Weekend Revenue for U.S. Box Office Movies, 2005-2007}
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\renewcommand{\thesubsection}{\thesection.\Alph{subsection}}
+
+\title{Modeling Opening Weekend Box Office Revenue for Domestic Movies, 2005-2007}
 
@@
 \section{Introduction}
-The project objective is to predict as accurately as possible opening weekend box office revenue for a set of movies, specifically from 2005-2007.
-Motivation for this project is a wealth of data points from both The Movie Database (TMDB) and Internet Movie Database (IMDb) around historical movies and TV shows: user-submitted ratings, applicable genres, budget, release date, actors, producer(s), production companies, country (or countries) of origin, etc.
-Further motivation is a prior understanding of likely useful predictors, such as the movie's budget, or the seasonal factors affecting a movie release.
-For example, the periods January-February, and to a lesser extent August-September, have been referred to as the ``dump months, '' i.e.
-effectively a dumping ground for low-potential movies that are not expected to make as much money.
-Various factors such as the seasonality of consumer spending (i.e binge before the holidays, pay down debt later), the timing of Golden Globes and Oscars awards, weather, the Super Bowl, and vacation months may all lead to seasonality in box office receipts.
-The years 2005-2007 were specifically chosen since they predate the rise of the streaming era and its potential impacts on box office sales.
-They are a short enough period that no inflation adjustment is made to any financial variables.
+The aim of this study was to explore what factors made films successful in attracting viewers to theater releases.
+Our primary motivation was personal interest as life-long “movie-buffs,” with extensive experience at the box office.
+Furthermore, we were encouraged by the wealth of free and publicly available data on films and their performance in theaters.
+To demarcate bounds, we decided to evaluate film releases prior to the rise of streaming—we felt it would be very difficult to account for the potential impacts of streaming on consumer behavior and the box office industry at large.
+In this vein—and to control for any shifting societal, political, economic or inflationary factors—we chose to focus on films released between 2005 and 2007.
+We chose to study feature-length films released in domestic theaters, filtering out short films, and keeping in line with our own experience.
+Finally, we decided to limit the timeframe of performance to the industry-standard “opening weekend,” allowing us to best control for variance in theater runs.
+Provided data availability, we determined revenue would be the best target variable to quantify viewership, assuming ticket prices were more or less equivalent across theaters. 
+Having reflected on our own experiences, we felt potentially important covariates would be centered around the “intrinsic” characteristics of each film: the business elements behind its production, its subject matter, its prequels (if any), its cast actors, and its hired crew.
 \section{Data}
+\subsection{Collection}
+We gathered our data from The Movie Database and Box Office Mojo.
+The Movie Database (“TMDB”) is a user-generated database of film and television, consisting of information and ratings on almost one million films internationally.
+Box Office Mojo is a subsidiary of IMDB which reports on domestic and international box-office statistics as far back as 1977.
+The Movie Database (“TMDB”) REST API was queried for English-language films distributed from 2005 to 2007, then cross-referenced with Box Office Mojo’s domestic opening-weekend statistics.
+The resulting initial dataset comprised 527 films, including data on: release date, opening weekend performance (e.g.
+revenue, theater distribution), film characteristics (e.g.
+genre, sequels), and production (e.g.
+budget, production companies).
+As a user-generated platform, TMDB also provided the community’s aggregate film ratings, along with the quantity of submitted votes.
+These latter features were initially considered as candidates for modeling but subsequently dropped, given the timing of user-submission couldn’t be controlled for.
+The TMDB API was then queried for supplemental information on the production, cast, and crew.
+This included features quantifying: the production headcount (e.g.
+size of writer dept., size of producing dept.), the experience of cast and crew (e.g.
+number of films the director had previously directed), the reputation of cast and crew (e.g.
+the median rating of films the writers had previously written), the size of the producing studio (e.g.
+independent or major), and if the screenplay was based on a prior work (e.g.
+novel).
+This exercise effectively doubled our features, with the aim to capture the human element of talent in film production.
 
-Data on approximately 500 movies that opened in U.S.
-theaters between 2005-2007 was aggregated from TMDB and IMDb.
-Only English-language movies are included.
-Retrieved data points include the release date, the opening weekend revenue, budget, and the number of theaters the movie opened in.
-A selection of genres that each movie could be classified as (e.g.
-drama, thriller, action, crime, documentary, etc.) is included.
-Most movies are tagged with two to three genres, though a handful go as high as six.
-Popularity, a proprietary TMDB metric, is also included.
-It itself is based on various data points including votes, views, release date, etc..
-As a community-built database, TMDB provides the data point vote average, namely the mean of user-submitted ratings on the scale of 0-10, as well as the number of submitted votes.
-Both the origin and production country (or countries) are included.
-The production companies are provided (e.g big hitters such as Paramount, Warner Bros, MGM).
-Finally, the runtime, and what collection the movie belongs to, if any, such as the Harry Potter series or the Pirates of the Caribbean collection.
-The operating revenue and number of opening theaters were specifically pulled from Box Office Mojo, a service of IMDB Pro.
-The following predictors are chosen for initial modeling:
-\begin{enumerate}
-\item Budget (continuous)
-\item Number of opening theaters (continuous)
-\item Runtime (continuous)
-\item Number of production companies (continuous)
-\item From a series (binary, 118 total)
-\item Opening weekend season (categorical)
-\begin{enumerate}
-\item Winter: January-February (71 total)
-\item Spring: March-April (84 total)
-\item Summer: May-August (154 total)
-\item Fall: September-October (120 total)
-\item Holiday: November-December (82 total)
-\end{enumerate}
-\end{enumerate}
+\subsection{Cleansing}
+According to the Screen Actors Guild\footnote{https://www.sagawards.org/awards/rules-eligibility/eligibility-criteria}, a feature-length film is defined at a minimum threshold of 60 minutes.
+Observations not meeting this runtime were dropped, along with films erroneously duplicated under different identifiers.
+In the feature space, we eliminated any irrelevant identifiers, and features beyond the scope of our observation.
+This included performance metrics recorded concurrently or after our target variable: revenue, opening weekend’s percent of total revenue, and opening theaters.
+This also included any user-generated features on TMDB mentioned above, including popularity and user ratings.
+Finally, the names of writers and directors were dropped, given the nominal features were far too granular to appropriately account for, instead opting to design aggregate features to proxy for experience and reputation.
+All missing values were checked against base data to confirm they weren’t development errors.
+Missing values deemed “reasonable,” i.e.
+indicating a lack of experience, were filled with a value of zero, and then proxied within a generated binary indicator.
+Missing values deemed collection errors were simply imputed with their median value.
+The primary nominal features were problematically encoded as a series of nested dictionaries within lists.
+These nominal features were painstakingly “unwrapped”, then one-hot encoded into binary columns for each level.
+Features of significance and reasonable size were left in this form; such as genre, which had anywhere from one to six assignments per film.
+Other features such as production studios, countries filmed in, and origin country were either consolidated into more reasonable sized groups, or transformed into more meaningful form.
+For example, films in the dataset had been shot in 32 countries; these countries were one-hot encoded, and grouped by continent.
+There were, however, far more unique production studios involved; these were classified as “major” or “independent” studios, then aggregated into a “percent\_independent” feature along with a binary indicator.
+The only exception to this exercise was the release date of the film, which we grouped into seasonal or monthly subsets.
+The output, ultimately, of this exercise was a 511-observation dataset, with 73 features describing the “intrinsic” components of film we first sought to characterize.
 
-Summary statistics are as follows:
+\subsection{Summary Statistics}
+Summary statistics for continuous features are as follows:
 \begin{table}[H]
 \centering
-\caption{Summary Statistics (n = 511)}
-\begin{tabular}{|lrrr|} 
-\hline
-Variable & Mean & Min & Max \\ 
-\hline
-opening revenue (\$) & 14,743,982 & 9,653 & 151,116,516\\
-\hline
-budget (\$) & 40,365,469 & 0 & 300,000,000\\
-\hline
-\# of theaters & 2,134 & 1 & 4,362\\
-\hline
-runtime (min) & 107 & 14 & 187\\
-\hline
-\# of prod. cos. & 3.8 & 0 & 17\\
-\hline
+\caption{Summary Statistics (n = 511)\label{summary_stats}}
+\begin{tabular}{lrrr}
+\toprule
+ & mean & min & max \\
+\midrule
+opening\_revenue (\$MM) & 14.75 & 0.01 & 151.12 \\
+budget (\$MM) & 42.88 & 0.00 & 300.00 \\
+runtime & 107.33 & 68.00 & 187.00 \\
+num\_prod\_cos & 1.56 & 0.00 & 7.00 \\
+previous\_film\_rating & 0.75 & 0.00 & 8.02 \\
+cast\_xp\_median & 26.45 & 0.00 & 93.00 \\
+cast\_rating\_max & 9.38 & 0.00 & 10.00 \\
+writer\_xp\_median & 11.26 & 0.00 & 136.00 \\
+writer\_xp\_sum & 27.87 & 0.00 & 366.00 \\
+writer\_rating\_median & 5.22 & 0.00 & 10.00 \\
+writer\_rating\_max & 6.57 & 0.00 & 10.00 \\
+production\_room & 16.71 & 1.00 & 100.00 \\
+writers\_room & 2.84 & 1.00 & 26.00 \\
+sound\_room & 13.59 & 1.00 & 70.00 \\
+crew & 96.05 & 2.00 & 573.00 \\
+\bottomrule
 \end{tabular}
@@
 
+Counts of films by release month and by genre are found in Tables \ref{month_distribution} and \ref{genre_count}, respectively.
+
 \section{Methods}
+\subsection{Baseline Model}
+We initially select budget (continuous), series (binary), number of production companies (continuous), runtime (continuous), and season (categorical) to establish a baseline model, with season categorized as Winter (Jan-Feb), Spring (Mar-Apr), Summer (May-Aug), Fall (Sep-Oct), and Holiday (Nov-Dec).
+The relationships between these predictors and opening revenue are visually examined.
+From pairs plots of the continuous predictors, budget and opening revenue seem positively, linearly related.
+There is no apparent relationship between opening revenue and the number of production companies or the runtime of the movie.
+No significant pairwise correlation appears between the predictors; each predictor displays wide variability, with budget and opening revenue both highly skewed right.
+Spring, summer, and holiday seasons appear to be associated with higher opening revenue on average, though this is skewed, especially by releases grossing over \$50M.
+Movies that are part of a series also appear associated with higher opening revenue on average.
+Initially, a model incorporating all variables produces $R^2_a$ around 0.6, but demonstrates clear heteroscedasticity of residuals.
+A square-root transformation noticeably stabilizes the variance, and establishes a baseline model, for which we perform residual diagnostics.
+Both DFFITS and DFBETAS analyses are considered to identify potentially influential observations.
+VIF analysis is also performed.
+
+\subsection{Alternate Models}
+We then attempt to improve upon the baseline by incorporating variables from our secondary set.
+We also consider interactions between different genres, and between months and genres.
+An automated approach taken is backwards elimination, including all interactions between genre pairs and between months and genres.
+Models are fit successively, dropping the least significant predictor.
+This process identifies a model achieving $R^2_a$ of 0.75, with noticeable heteroscedasticity and very high multicollinearity, and is not considered further.
+In parallel, we examine the effect or non-effect of adding individual predictors or classes of predictors, including an exhaustive analysis of the “indie” status of a film, and considering scenarios where indie and non-indie production companies collaborated.
 
 \section{Results}
+After scaling budget and opening revenue to millions of dollars for easier interpretation, the baseline model specification is:
+
+\begin{table}[H]
+\centering
+  \begin{threeparttable}
+   \scriptsize
+    \caption{Baseline Model}
+	\begin{tabular}{lrrrrr}
+	\toprule
+	                                    & coef & t & P>|t| & 0.025 & 0.975 \\
+	\midrule
+	const          &       2.9980       &     6.799  &         0.000        &        2.132    &        3.864     \\
+	budget         &       0.0330       &    18.651  &         0.000        &        0.030    &        0.036     \\
+	is\_series     &       1.2225       &     5.974  &         0.000        &        0.820    &        1.624     \\
+	num\_prod\_cos &      -0.2793       &    -3.905  &         0.000        &       -0.420    &       -0.139     \\
+	runtime        &      -0.0027       &    -0.652  &         0.514        &       -0.011    &        0.005     \\
+	is\_spring     &      -0.3056      &    -1.328  &         0.185        &       -0.758    &        0.147     \\
+	is\_summer     &      -0.4470       &    -2.143  &         0.033        &       -0.857    &       -0.037     \\
+	is\_fall       &      -1.0822      &    -5.008  &         0.000        &       -1.507    &       -0.658     \\
+	is\_holiday    &      -1.0853       &    -4.587  &         0.000        &       -1.550    &       -0.620     \\
+	\bottomrule
+	\end{tabular}
+    \begin{tablenotes}
+      \scriptsize
+      \item $R^2_a$ = 0.54
+    \end{tablenotes}
+  \end{threeparttable}    
+\end{table}
+
+An increase of \$1 million in the film’s budget is associated with an expected increase of \$33 thousand in the square root of opening revenue, all else held equal.
+Movies that belong to a series are associated with higher opening weekend revenue, with an estimated increase (in the square root) of \$1.22 million.
+We do not have significant evidence for runtime influencing opening revenue, while the number of production companies of a movie is associated with an expected decrease in opening revenue.
+Our reference group for the categorical covariate season is Winter, for which the expected (square root) opening revenue is \$3 million.
+Surprisingly, expected (square root) opening revenue is lower for all the remaining seasons, dropping to \$2.7 million for Spring, \$2.6 million for Summer, and \$1.9 million for the Fall and Holiday seasons.
+
+Residual diagnostics (Figures \ref{baseline_residuals_vs_fitted}-\ref{baseline_residuals_vs_leverage}) show some heteroscedasticity, which the square-root transformation had improved but not completely remedied.
+Otherwise, the residuals appear normally distributed, and there appear to be no glaring instances of influential observations.
+
+A DFFITS analysis (Table \ref{dffits}) identifies 32 observations possibly worth investigating.
+Performing the same regression while dropping these does not significantly change the coefficients, and all are ultimately retained.
+Since a full DFBETAS analysis is prohibitive, we re-estimate the model after dropping the top observation affecting the estimation of the budget coefficient (Table \ref{dfbetas_budget}), and similarly for the is\_series coefficient (Table \ref{dfbetas_series}).
+Neither coefficient changes materially, so we retain all observations.
+Finally, the average VIF is 7, which does not warrant remediation.
+
+Our final model uses forward stepwise selection with continuous covariates, including variables from the secondary set around crew, cast, and director.
+We create a saturated model, then identify the final retained interactions with backwards elimination and forward selection.
+For genre-month interaction, we include horror movies released in October and family movies relased in December.
+For genre-genre interactions, we include rom-com and family-animation effects.
+Finally, we include a series and sci-fi effect.
+The final model specification is:
+
+\begin{table}[H]
+\centering
+  \begin{threeparttable}
+  \scriptsize
+    \caption{Final Model}
+	\begin{tabular}{lrrrrrr}
+	\toprule
+	                                    & coef & t & P>|t| & 0.025 & 0.975 \\
+		\midrule
+		const                        &       2.5303       &     8.417  &         0.000        &        1.940    &        3.121     \\
+		budget                       &       0.0298       &    15.937  &         0.000        &        0.026    &        0.033     \\
+		num\_prod\_cos   &      -0.2908       &    -4.255  &         0.000        &       -0.425    &       -0.156     \\
+		crew                         &       0.0034       &     4.268  &         0.000        &        0.002    &        0.005     \\
+		cast\_xp\_median             &      -0.0089       &    -2.349  &         0.019        &       -0.016    &       -0.001     \\
+		is\_series                   &      -2.2463       &    -1.325  &         0.186        &       -5.579    &        1.086     \\
+		previous\_film\_release      &      -0.0003       &    -2.909  &         0.004        &       -0.001    &       -0.000     \\
+		previous\_film\_rating       &       0.5557      &     2.205  &         0.028        &        0.060    &        1.051     \\
+		first\_time\_directors       &      -0.7159       &    -2.649  &         0.008        &       -1.247    &       -0.185     \\
+		is\_feb                      &       0.3944       &     1.251  &         0.212        &       -0.225    &        1.014     \\
+		is\_mar                      &       0.3185       &     1.021  &         0.308        &       -0.295    &        0.932     \\
+		is\_apr                      &      -0.2979       &    -0.921  &         0.357        &       -0.933    &        0.337     \\
+		is\_may                      &      -0.3588       &    -1.038  &         0.300        &       -1.038    &        0.321     \\
+		is\_jun                      &       0.1060       &     0.329  &         0.743        &       -0.528    &        0.739     \\
+		is\_jul                      &      -0.4237       &    -1.269  &         0.205        &       -1.079    &        0.232     \\
+		is\_aug                      &       0.0965       &     0.306  &         0.760        &       -0.524    &        0.717     \\
+		is\_sep                      &      -0.7991       &    -2.730  &         0.007        &       -1.374    &       -0.224     \\
+		is\_oct                      &      -0.6188       &    -1.925  &         0.055        &       -1.251    &        0.013     \\
+		is\_nov                      &      -0.1301       &    -0.396  &         0.692        &       -0.775    &        0.515     \\
+		is\_dec                      &      -1.2243       &    -3.671  &         0.000        &       -1.880    &       -0.569     \\
+		is\_horror                   &       0.4524      &     2.061  &         0.040        &        0.021    &        0.884     \\
+		is\_romance                  &      -0.6347       &    -2.596  &         0.010        &       -1.115    &       -0.154     \\
+		is\_comedy                   &       0.0364       &     0.226  &         0.821        &       -0.280    &        0.352     \\
+		is\_family                   &      -0.1595       &    -0.780  &         0.436        &       -0.561    &        0.242     \\
+		is\_animation                &      -0.3852       &    -0.586  &         0.558        &       -1.676    &        0.906     \\
+		is\_sci\_fi                  &      -0.5327       &    -2.122  &         0.034        &       -1.026    &       -0.039     \\
+		is\_drama                    &      -0.2956       &    -2.101  &         0.036        &       -0.572    &       -0.019     \\
+		is\_oct\_*\_is\_horror       &       1.3565       &     2.474  &         0.014        &        0.279    &        2.434     \\
+		is\_dec\_*\_is\_family       &       0.9714       &     1.959  &         0.051        &       -0.003    &        1.946     \\
+		is\_comedy\_*\_is\_romance   &       1.2247       &     3.944  &         0.000        &        0.615    &        1.835     \\
+		is\_animation\_*\_is\_family &       1.1228       &     1.541  &         0.124        &       -0.309    &        2.555     \\
+		is\_series\_*\_is\_sci\_fi   &       1.3973       &     2.423  &         0.016        &        0.264    &        2.531     \\
+		\bottomrule
+		\end{tabular}
+    \begin{tablenotes}
+      \scriptsize
+      \item $R^2_a$ = 0.63
+    \end{tablenotes}  
+  \end{threeparttable}    
+\end{table}
+
+Residual diagnostics (Figures \ref{final_residuals_vs_fitted}-\ref{final_residuals_vs_leverage}) continue to show some heteroscedasticity, but otherwise appear normally distributed with no notable influential observations.
+Multicollinearity is slighter worse than the baseline model, with mean VIF around 9.
+Budget and number of prediction companies have similar effects as in the baseline model, and monthly effects generally mirror the seasonality in the baseline model.
 
 \section{Discussion}
+Various modeling tradeoffs were encountered between the baseline model and alternate models.
+Creating the categorical variable season led to improved residual diagnostics, while treating each month individually led to more significant interactions.
+“Indie” status was not significant, either initially modeling it as a binary variable, or treating it continously to capture scenarios where both major and indie studios collaborated on a film, and indie films could be considered as a percentage of total studio contribution.
+We had expected lower revenue on average in the Winter months, but found the opposite.
+These have been referred to as the “dump months” in the film industry, effectively a dumping ground for low-potential movies, so our finding was unexpected.
 
-\bibliographystyle{sdss2020} % Please do not change the bibliography style
-%\bibliography{fullbib}
+The modeling approach taken (i.e. initially posited predictors vs automated selection) also demonstrated tradeoffs. Both models confirmed budget as an important predictor of opening revenue.
+Automated selection led to an increase in $R^2_a$, but tended to worsen heteroscedasticity and introduce multicollinearity.
+Cross-validation would have paired well with automated selection methods, leading to replicatable results.
+
+Finally, the use of only classical regression methods was somewhat limiting, for example in initially stabilizing the variance.
+Further possible avenues of exploration include implementation of cross-validation, additional feature cleansing or engineering, additional data such as popularity metrics or revenue by region, and implementation of modern regression techniques.
+
+%\bibliographystyle{sdss2020} 
+%\bibliography{bibl}
+
+\section*{References}
+
+\flushleft{TMDB data sourced using the TMDB REST API, version 3. The version 3 API documentation can be found at \mbox{https://developer.themoviedb.org/docs/getting-started}.}\\[1em]
+\flushleft{Box Office Mojo. (n.d.). \textit{Domestic Box Office For 2005}. \mbox{https://www.boxofficemojo.com/year/2005}}\\[1em]
+\flushleft{Box Office Mojo. (n.d.). \textit{Domestic Box Office For 2006}. \mbox{https://www.boxofficemojo.com/year/2006}}\\[1em]
+\flushleft{Box Office Mojo. (n.d.). \textit{Domestic Box Office For 2007}. \mbox{https://www.boxofficemojo.com/year/2007}}\\[1em]
 
 \appendix
+\clearpage
+
+\section{Supplemental Visualizations}
+\begin{figure}[H]
+\centering
+	\centerline{\includegraphics[width=\columnwidth]{baseline_residuals_vs_fitted}}
+	\caption{Residuals vs. Fitted Plot, Baseline Model\label{baseline_residuals_vs_fitted}}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\centerline{\includegraphics[width=\columnwidth]{baseline_qq}}
+	\caption{Q-Q Plot, Baseline Model\label{baseline_qq}}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\centerline{\includegraphics[width=\columnwidth]{baseline_scale_location}}
+	\caption{Scale-Location Plot, Baseline Model\label{baseline_scale_location}}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\centerline{\includegraphics[width=\columnwidth]{baseline_residuals_vs_leverage}}
+	\caption{Residuals vs. Leverage Plot, Baseline Model\label{baseline_residuals_vs_leverage}}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\centerline{\includegraphics[width=\columnwidth]{final_residuals_vs_fitted}}
+	\caption{Residuals vs. Fitted Plot, Final Model\label{final_residuals_vs_fitted}}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\centerline{\includegraphics[width=\columnwidth]{final_qq}}
+	\caption{Q-Q Plot, Final Model\label{final_qq}}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\centerline{\includegraphics[width=\columnwidth]{final_scale_location}}
+	\caption{Scale-Location Plot, Final Model\label{final_scale_location}}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\centerline{\includegraphics[width=\columnwidth]{final_residuals_vs_leverage}}
+	\caption{Residuals vs. Leverage Plot, Final Model\label{final_residuals_vs_leverage}}
+\end{figure}
+
+\section{Supplemental Analyses}
+
+\begin{table}[H]
+\centering
+\caption{Distributions by month of release\label{month_distribution}}
+\begin{tabular}{lr}
+\toprule
+month & count\\
+\midrule
+is\_jan & 28 \\
+is\_feb & 42 \\
+is\_mar & 45 \\
+is\_apr & 39 \\
+is\_may & 31 \\
+is\_jun & 43 \\
+is\_jul & 35 \\
+is\_aug & 45 \\
+is\_sep & 71 \\
+is\_oct & 50 \\
+is\_nov & 37 \\
+is\_dec & 45 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\centering
+\caption{Count of genres\label{genre_count}}
+\begin{tabular}{lr}
+\toprule
+genre & count  \\
+\midrule
+is\_adventure & 95 \\
+is\_fantasy & 51 \\
+is\_animation & 33 \\
+is\_drama & 240 \\
+is\_horror & 59 \\
+is\_action & 103 \\
+is\_comedy & 223 \\
+is\_history & 25 \\
+is\_thriller & 132 \\
+is\_crime & 86 \\
+is\_documentary & 11 \\
+is\_sci\_fi & 39 \\
+is\_mystery & 53 \\
+is\_music & 17 \\
+is\_romance & 107 \\
+is\_family & 87 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\caption{DFFITS results, baseline model (Top 10)\label{dffits}}
+\begin{tabular}{llr}
+\toprule
+ obs & release & dffits \\
+\midrule
+282 & Superman Returns & -0.56 \\
+438 & Pirates of the Caribbean: At World's End & -0.53 \\
+465 & The Bourne Ultimatum & 0.52 \\
+380 & Evan Almighty & -0.50 \\
+47 & Harry Potter and the Goblet of Fire & 0.47 \\
+102 & Son of the Mask & -0.46 \\
+178 & Basic Instinct 2 & -0.45 \\
+238 & Letters from Iwo Jima & -0.45 \\
+103 & Star Wars: Episode III - Revenge of the Sith & 0.44 \\
+339 & 300 & 0.39 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\caption{DFBETAs for budget, baseline model (Top 10)\label{dfbetas_budget}}
+\begin{tabular}[t]{l L{0.4\linewidth} rr}
+%\begin{tabular}{llr}
+\toprule
+obs & release & dfbetas & budget (\$MM) \\
+\midrule
+438 & Pirates of the Caribbean: At World's End & -0.37 & 300.00 \\
+380 & Evan Almighty & -0.33 & 175.00 \\
+282 & Superman Returns & -0.31 & 223.00 \\
+260 & Poseidon & -0.28 & 160.00 \\
+401 & I Am Legend & 0.24 & 150.00 \\
+419 & Meet the Robinsons & -0.24 & 150.00 \\
+238 & Letters from Iwo Jima & 0.23 & 19.00 \\
+212 & Flushed Away & -0.23 & 149.00 \\
+472 & The Golden Compass & -0.22 & 180.00 \\
+452 & Shrek the Third & 0.21 & 160.00 \\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+\caption{DFBETAs for is\_series, baseline model (Top 10)\label{dfbetas_series}}
+\begin{tabular}[t]{l L{0.5\linewidth} rr}
+\toprule
+obs & release & dfbetas & is\_series \\
+\midrule
+465 & The Bourne Ultimatum & 0.33 & 1 \\
+238 & Letters from Iwo Jima & -0.31 & 1 \\
+267 & Saw III & 0.29 & 1 \\
+102 & Son of the Mask & -0.29 & 1 \\
+103 & Star Wars: Episode III - Revenge of the Sith & 0.29 & 1 \\
+97 & Saw II & 0.28 & 1 \\
+178 & Basic Instinct 2 & -0.26 & 1 \\
+449 & Saw IV & 0.26 & 1 \\
+47 & Harry Potter and the Goblet of Fire & 0.25 & 1 \\
+230 & Jackass Number Two & 0.23 & 1 \\
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 \label{a:code}
-Data and code are available at https://github.com/cons-code/MATH-5500
+Replication data and code can be found in the GitHub repo \mbox{https://github.com/cons-code/MATH-5500}.
 \end{document}

### 2024-11-30 Final/Baseline_Hockey.Rmd
Baseline Model

---
title: "R.Final_PenalizedRegression"
author: "Connor Logue"
date: "2024-11-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Data
```{r}
data = read.csv("data.csv", header = TRUE)
data
```

### EDA  & Cleaning
```{r}
paste("Number of Instances:", nrow(data))
paste("Number of Variables:", ncol(data))
```

```{r}
head (data)
```

### Baseline
```{r}
base <- lm(Salary_log ~ FO_pct + G + A + GF + GA + plus_mins + TOI + iSF + Yrs_in_NHL + Ht + Wt + Grit + iHF + iHA + iBLK + PIM + GP, data = data)
```

``` {r}
summary(base)
```

```{r}
vif(base)
mean(vif(base))
```

```{r}
RSS <- c(crossprod(base$residuals))
MSE <- RSS / length(base$residuals)
MSE
```

```{r}
base2 <- lm(Salary_log ~ G + A + GF + GA + plus_mins + TOI + iSF + Yrs_in_NHL + GP, data = data)
summary(base2)
```
```{r}
vif(base2)
mean(vif(base2))
```

### 2024-12-01 Final/Baseline_Penalized.Rmd
Bedrock, Elastic Model Selection, Baseline, Ridge Shrinkage

---
title: "Con"
author: "Connor Logue"
date: "2024-12-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries
```{r}
library(dplyr)
library(glmnet)
library(car)
```

### Data
```{r}
data = read.csv("combined.csv", header = TRUE)
data
```

### Cleansing
```{r}
data$Yrs_in_NHL <- 2016 - data$DftYr
data$A = data$A1 + data$A2

data$GF[is.na(data$GF)] <- median(data$GF, na.rm=TRUE)
data$GA[is.na(data$GA)] <- median(data$GA, na.rm=TRUE)
data$iSF[is.na(data$iSF)] <- median(data$iSF, na.rm=TRUE)
data$Yrs_in_NHL[is.na(data$Yrs_in_NHL)] <- 0
data$iHA[is.na(data$iHA)] <- median(data$iHA, na.rm=TRUE)

data <- data[data$Yrs_in_NHL > 3,]
data <- data[data$Is_Offense == 1,]

scale_numeric <- function(x) x %>% mutate_if(is.numeric, function(y) as.vector(scale(y)))
scaled_data <- data %>% scale_numeric()
```

```{r}
bedrock <- lm(Salary_log ~ FO_pct + G + A + GF + GA + plus_mins + TOI + iSF + Yrs_in_NHL + Ht + Wt + Grit + iHF + iHA + iBLK + PIM + GP, data = scaled_data)
summary(bedrock)

e     <- resid(bedrock)
Yhat  <- predict(bedrock)
s     <- summary(bedrock)$sigma
es    <- e/s
```

```{r}
plot(Yhat, e, type = 'n', main = 'Fitted vs Residual')
grid()
abline(h = 0, lwd = 2, col = 'gray')
points(Yhat, e, pch = 16, col = 'darkblue')
```

```{r}
qq_model  <- qqnorm(es, plot.it = FALSE)
qqplot(qq_model$x, qq_model$y, xlab = 'E(e*)', ylab = 'e*', pch = 16, col = 'darkblue',
       main = 'Q-Q Plot')
qqline(qq_model$y, col = 'gray', lwd = 2)
grid()
```

```{r}
vif(bedrock)
mean(vif(bedrock))
```
```{r, fig.width=10, fig.height=10}
pairwise <- subset(scaled_data, select = c(FO_pct, G, A, GF, GA, plus_mins, TOI, iSF, Yrs_in_NHL, Ht, Wt, Grit, iHF, iHA, iBLK, PIM, GP))

library(corrplot)
M = cor(pairwise)
corrplot(M, method = 'number')
```

### Penalized Regression
``` {r}
library(glmnet)
X         <- model.matrix(Salary_log ~ FO_pct + G + A + GF + GA + plus_mins + TOI + iSF + Yrs_in_NHL + Ht + Wt + Grit + iHF + iHA + iBLK + PIM + GP -1, data = scaled_data)
Y         <- scaled_data$Salary_log

ridge_cv   <- cv.glmnet(x = X, y = Y, family = "gaussian", alpha = 0, type.measure = "mse", nfolds = nrow(scaled_data), grouped = FALSE)
elastic_cv <- cv.glmnet(x = X, y = Y, family = "gaussian", alpha = 0.5, type.measure = "mse", nfolds = nrow(scaled_data), grouped = FALSE)
lasso_like_cv <- cv.glmnet(x = X, y = Y, family = "gaussian", alpha = 0.9, type.measure = "mse", nfolds = nrow(scaled_data), grouped = FALSE)
lasso_cv <- cv.glmnet(x = X, y = Y, family = "gaussian", alpha = 1, type.measure = "mse", nfolds = nrow(scaled_data), grouped = FALSE)

results <- matrix(c(ridge_cv$lambda[which.min(ridge_cv$cvm)], min(ridge_cv$cvm),
              elastic_cv$lambda[which.min(elastic_cv$cvm)],min(elastic_cv$cvm),
              lasso_like_cv$lambda[which.min(lasso_like_cv$cvm)], min(lasso_like_cv$cvm),
              lasso_cv$lambda[which.min(lasso_cv$cvm)], min(lasso_cv$cvm)), 
            nrow = 4, ncol = 2, byrow = TRUE)

rownames(results) = c("Ridge", "Elastic", "Lasso-Like", "Lasso")
colnames(results) = c("Lambda", "MSE")
results
```
```{r}
plot(elastic_cv)
```

```{r}
coef(elastic_cv)
```

```{r}
baseline <- lm(Salary_log ~ G + A + GF + plus_mins + TOI + iSF + Yrs_in_NHL + Wt + GP, data = scaled_data)
summary(baseline)

e     <- resid(baseline)
Yhat  <- predict(baseline)
s     <- summary(baseline)$sigma
es    <- e/s
```

```{r}
plot(Yhat, e, type = 'n', main = 'Fitted vs Residual')
grid()
abline(h = 0, lwd = 2, col = 'gray')
points(Yhat, e, pch = 16, col = 'darkblue')
```

```{r}
qq_model  <- qqnorm(es, plot.it = FALSE)
qqplot(qq_model$x, qq_model$y, xlab = 'E(e*)', ylab = 'e*', pch = 16, col = 'darkblue',
       main = 'Q-Q Plot')
qqline(qq_model$y, col = 'gray', lwd = 2)
grid()
```

```{r}
vif(baseline)
mean(vif(baseline))
```

```{r}
library(glmnet)
X         <- model.matrix(Salary_log ~ G + A + GF + plus_mins + TOI + iSF + Yrs_in_NHL + Wt + GP -1, data = scaled_data)
Y         <- scaled_data$Salary_log

ridge_cv   <- cv.glmnet(x = X, y = Y, family = "gaussian", alpha = 0, type.measure = "mse", nfolds = nrow(scaled_data), grouped = FALSE)

plot(ridge_cv)
```

```{r}
coef(ridge_cv)
```

### 2024-12-16 Final/R/generalized_least_squares.R
initial commit of gls module

@@ -1,0 +1,223 @@
+### FINAL PROJECT: GLS ###
+
+## LIBRARY INSTALLATION ##
+
+library(car)
+library(dplyr)
+library(nlme)
+library(glmnet)
+library(xtable)
+
+
+## DATA PROCESSING ##
+
+# Read CV containing combined NHL dataframe.
+nhl_df <- read.csv("./Georgetown/MATH 5500/final_project/combined.csv", 
+                   header=TRUE)
+
+# Perform data cleaning.
+nhl_df$Yrs_in_NHL <- 2016 - nhl_df$DftYr
+nhl_df$A          <- nhl_df$A1 + nhl_df$A2
+
+nhl_df$GF[is.na(nhl_df$GF)]   <- median(nhl_df$GF, na.rm=TRUE)
+nhl_df$GA[is.na(nhl_df$GA)]   <- median(nhl_df$GA, na.rm=TRUE)
+
+nhl_df$iSF[is.na(nhl_df$iSF)] <- median(nhl_df$iSF, na.rm=TRUE)
+nhl_df$iHA[is.na(nhl_df$iHA)] <- median(nhl_df$iHA, na.rm=TRUE)
+
+nhl_df$Yrs_in_NHL[is.na(nhl_df$Yrs_in_NHL)] <- 0
+
+nhl_df <- nhl_df[nhl_df$Yrs_in_NHL > 3,]
+nhl_df <- nhl_df[nhl_df$Is_Offense == 1,]
+
+scale_numeric <- function(x) x %>% mutate_if(is.numeric, function(y) as.vector(scale(y)))
+nhl_df_scaled <- nhl_df %>% scale_numeric()
+
+# Extract baseline parameters from scaled NHL dataframe.
+nhl_df_base <- data.frame(Salary_log=nhl_df_scaled$Salary_log,
+                          FO_pct=nhl_df_scaled$FO_pct,
+                          G=nhl_df_scaled$G,
+                          A=nhl_df_scaled$A,
+                          GF=nhl_df_scaled$GF,
+                          GA=nhl_df_scaled$GA,
+                          plus_mins=nhl_df_scaled$plus_mins,
+                          TOI=nhl_df_scaled$TOI,
+                          iSF=nhl_df_scaled$iSF,
+                          Yrs_in_NHL=nhl_df_scaled$Yrs_in_NHL,
+                          Ht=nhl_df_scaled$Ht,
+                          Wt=nhl_df_scaled$Wt,
+                          Grit=nhl_df_scaled$Grit,
[…]

### 2024-12-16 Final/Reg-Final_12.12.2000.Rmd
Squashed commit of the following:
commit 68f8653b09ea6801b0fbe145e60b4e10a5c68dfa
Author: mernaghp <bigbenshuffleboard@gmail.com>
Date:   Mon Dec 16 17:22:09 2024 -0500

    Paper minor edits

commit c97b16e7507fe799428ff684a2bc597bf43c5156
Author: mernaghp <bigbenshuffleboard@gmail.com>
Date:   Mon Dec 16 15:47:04 2024 -0500

    More paper revisions

commit ebd2ccd875820546915c30fee9b15df9ca2fabeb
Merge: 95fd5b2 1136e8b
Author: mernaghp <bigbenshuffleboard@gmail.com>
Date:   Mon Dec 16 13:41:09 2024 -0500

    Merge branch 'main' into pm

commit 95fd5b2f444f19850552e204951e69ab4fd56681
Autho

---
title: "Con"
author: "Connor Logue"
date: "2024-12-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Libraries
```{r}
library(dplyr)
library(glmnet)
library(car)
library(nestedcv)
library(xtable)
library(corrplot)
```

### Data
```{r}
data = read.csv("combined.csv", header = TRUE)
data
```

### Cleansing
```{r}
data$Yrs_in_NHL <- 2016 - data$DftYr
data$A = data$A1 + data$A2
data$Salary_log_corrected <- data$Salary_log

data$GF[is.na(data$GF)] <- median(data$GF, na.rm=TRUE)
data$GA[is.na(data$GA)] <- median(data$GA, na.rm=TRUE)
data$iSF[is.na(data$iSF)] <- median(data$iSF, na.rm=TRUE)
data$Yrs_in_NHL[is.na(data$Yrs_in_NHL)] <- 0
data$iHA[is.na(data$iHA)] <- median(data$iHA, na.rm=TRUE)

data <- data[data$Yrs_in_NHL > 3,]
data <- data[data$Is_Offense == 1,]
```

### EDA | Pairwise Plots
```{r, fig.height=7.5}
plot(data$TOI, data$Salary_log)
lines(lowess(data$TOI, data$Salary_log), col=2)
```

### OLS
```{r}
data$TOI_c   <- scale(data$TOI, scale = FALSE)
data$TOI_c2  <- data$TOI_c^2
bedrock <- lm(Salary_log ~ G + A + GF + GA + plus_mins + TOI_c + TOI_c2 + iSF + Yrs_in_NHL + Ht + Wt + iHF + iHA + iBLK + PIM + GP, data = data)

summary(bedrock)
mean(vif(bedrock))
```

```{r}
vif(bedrock)
mean(vif(bedrock))

```

```{r, fig.height = 10}
par(mfrow = c(2,2))
plot(bedrock)
```

### Print
```{r}
#xtable(summary(bedrock_poly), digits = 3)
```

```{r}
#coef <- data.frame(coef(bedrock_poly))
#coef$expo_coef = (exp(coef$coef.bedrock_poly) - 1) * 1000000
#coef_sub <- subset(coef, select = expo_coef)
#xtable(coef_sub, digits = 0)
```

```{r}
#confidence <- data.frame(confint(bedrock_poly, level = 0.95))
#confidence$expo_lower = (exp(confidence$X2.5..) - 1) * 1000000
#confidence$expo_upper = (exp(confidence$X97.5..) - 1) * 1000000
#conf_sub <- subset(confidence, select = c(expo_lower, expo_upper))
#xtable(conf_sub, digits = 0)
```

### EDA | Simple Linear Regression
```{r}
uni <- lm(Salary_log ~ GP, data = data)
summary(uni)
```

### EDA | Correlation Plot
```{r, fig.width=10, fig.height=10}
pairwise <- subset(data, select = c(G, A, GF, GA, plus_mins, TOI, iSF, Yrs_in_NHL, Ht, Wt, iHF, iHA, iBLK, PIM, GP))
M = cor(pairwise)
corrplot(M, method = 'number')
```

### Elastic Net
```{R}
X         <- model.matrix(Salary_log ~ G + A + GF + GA + plus_mins + Yrs_in_NHL + TOI_c + TOI_c2 + iSF + Ht + Wt + iHF + iHA + iBLK + Grit + GP -1, data = data)
Y         <- data$Salary_log

elastic   <- glmnet(x = X, y = Y, family = "gaussian", intercept = TRUE, alpha = 0.5)
plot(elastic, xvar = "lambda", label = TRUE)
```

```{r}
elastic_cv   <- cv.glmnet(x = X, y = Y, family = "gaussian", alpha = 0.5, ype.measure = "mse", nfolds = nrow(data), grouped = FALSE)
coef(elastic_cv) 
```

```{r}
baseline <- lm(Salary_log ~ A + GF + plus_mins + Yrs_in_NHL + TOI_c + TOI_c2 +  iSF + Wt + iHF + iBLK + GP, data = data)
summary(baseline)
```

```{r}
vif(baseline)
mean(vif(baseline))
```

### Cross-Nested Validation
```{r}
library(glmnet)
library(nestedcv)
X         <- model.matrix(Salary_log ~ A + GF + plus_mins + TOI + iSF + Yrs_in_NHL + Wt + GP -1, data = scaled_data)
Y         <- scaled_data$Salary_log

test <- nestcv.glmnet(x = X, y = Y, family = "gaussian", alphaSet = seq(0, 0.3, 0.1), type.measure = "mse", n_outer_folds = 10, n_inner_folds = 10, grouped = FALSE)

plot(test$outer_result[[1]]$cvafit)

```
```{R}
summary(test)
```

### Penalized Regression

```{R}

group1 <- rep(1:20, each=5)
ml <- gglasso(x=bardet$x, y=bardet$y, group=group1, loss="ls")

```

```{R}

#X         <- model.matrix(Salary_log ~ G + A + GF + GA + plus_mins + TOI + TOI_2 + iSF + Yrs_in_NHL + Ht + Wt + iHF + iHA + iBLK + PIM + GP -1, data = scaled_data)
X <- subset(scaled_data, select = c(G, A, GF, GA ))
Y         <- scaled_data$Salary_log

group1 <- rep(1:2, each = 2)

fit_ls <- gglasso(x = test$X, y = test$Y, group = group1, loss = "ls")
```

```{r}
library(psych)
describe(data$Salary)
```
```{r}
hist(data$Salary)
```

```{r, fig.height=10}
plot(EDA[, 1:7])
```

### 2024-12-16 Final/Regression_Final_12.11.Rmd
Squashed commit of the following:
commit 68f8653b09ea6801b0fbe145e60b4e10a5c68dfa
Author: mernaghp <bigbenshuffleboard@gmail.com>
Date:   Mon Dec 16 17:22:09 2024 -0500

    Paper minor edits

commit c97b16e7507fe799428ff684a2bc597bf43c5156
Author: mernaghp <bigbenshuffleboard@gmail.com>
Date:   Mon Dec 16 15:47:04 2024 -0500

    More paper revisions

commit ebd2ccd875820546915c30fee9b15df9ca2fabeb
Merge: 95fd5b2 1136e8b
Author: mernaghp <bigbenshuffleboard@gmail.com>
Date:   Mon Dec 16 13:41:09 2024 -0500

    Merge branch 'main' into pm

commit 95fd5b2f444f19850552e204951e69ab4fd56681
Autho

--- 2024-12-01 Final/Baseline_Penalized.Rmd
@@
 library(car)
+library(nestedcv)
 ```
@@
 data$A = data$A1 + data$A2
+data$Salary_log_corrected <- data$Salary_log
 
@@
 scaled_data <- data %>% scale_numeric()
+
+EDA = subset(scaled_data, select = c(Salary_log, G, A, plus_mins, TOI, iSF, Yrs_in_NHL, Ht, Wt, Grit, iHF, iHA, iBLK, GP))
 ```
@@
 ```{r}
-bedrock <- lm(Salary_log ~ FO_pct + G + A + GF + GA + plus_mins + TOI + iSF + Yrs_in_NHL + Ht + Wt + Grit + iHF + iHA + iBLK + PIM + GP, data = scaled_data)
+library(psych)
+describe(data$Salary)
+```
+```{r}
+hist(data$Salary)
+```
+
+```{r, fig.height=10}
+plot(EDA[, 1:7])
+```
+
+```{r}
+bedrock <- lm(Salary_log ~ FO_pct + G + A + GF + GA + plus_mins + TOI + iSF + Yrs_in_NHL + Ht + Wt + iHF + iHA + iBLK + Grit + GP, data = scaled_data)
 summary(bedrock)
+```
+```{r}
+vif(bedrock)
+mean(vif(bedrock))
+```
 
-e     <- resid(bedrock)
-Yhat  <- predict(bedrock)
-s     <- summary(bedrock)$sigma
-es    <- e/s
+```{r, fig.height=10}
+par(mfrow = c(2,2))
+plot(bedrock)
+```
+
+### EDA
+```{r, fig.height=7.5}
+plot(scaled_data$TOI, scaled_data$Salary_log)
+lines(lowess(scaled_data$TOI, scaled_data$Salary_log), col=2)
+```
+
+```{r}
+scaled_data$A_2  <- scaled_data$A^2
+scaled_data$TOI_2  <- scaled_data$TOI^2
+
+scaled_data$iBLK_2  <- scaled_data$iBLK^2
+scaled_data$Grit_2  <- scaled_data$Grit^2
+
+bedrock_poly <- lm(Salary_log ~ G + A + GF + GA + plus_mins + TOI_2 + iSF + Yrs_in_NHL + Ht + Wt + iHF + iHA + iBLK + Grit + GP, data = scaled_data)
+
+summary(bedrock_poly)
+mean(vif(bedrock_poly))
+
+```
+
+```{r}
+vif(bedrock_poly)
+mean(vif(bedrock_poly))
+
+```
+
+```{r, fig.height = 10}
+par(mfrow = c(2,2))
+plot(bedrock_poly)
+```
+
+````{r}
+confint(bedrock_poly, level = 0.95)
+```
+
+```{R}
+library(xtable)
+conf = confint(bedrock_poly, level = 0.99)
+
+xtable(confint(bedrock_poly, level = 0.95))
+
 ```
@@
 
-```{r}
-vif(bedrock)
-mean(vif(bedrock))
-```
 ```{r, fig.width=10, fig.height=10}
-pairwise <- subset(scaled_data, select = c(FO_pct, G, A, GF, GA, plus_mins, TOI, iSF, Yrs_in_NHL, Ht, Wt, Grit, iHF, iHA, iBLK, PIM, GP))
+pairwise <- subset(scaled_data, select = c(G, A, GF, GA, plus_mins, TOI, iSF, Yrs_in_NHL, Ht, Wt, iHF, iHA, iBLK, Grit, GP))
 
@@
 library(glmnet)
-X         <- model.matrix(Salary_log ~ FO_pct + G + A + GF + GA + plus_mins + TOI + iSF + Yrs_in_NHL + Ht + Wt + Grit + iHF + iHA + iBLK + PIM + GP -1, data = scaled_data)
+X         <- model.matrix(Salary_log ~ G + A + GF + GA + plus_mins + TOI + iSF + Yrs_in_NHL + Ht + Wt + iHF + iHA + iBLK + Grit + GP -1, data = scaled_data)
 Y         <- scaled_data$Salary_log
@@
 ```
+```{r}
+summary(elastic_cv)
+```
+
+```{r}
+coef(elastic_cv, s = "lambda.min")
+```
 
@@
 ```{r}
-baseline <- lm(Salary_log ~ G + A + GF + plus_mins + TOI + iSF + Yrs_in_NHL + Wt + GP, data = scaled_data)
+baseline <- lm(Salary_log ~ A + GF + plus_mins + TOI + iSF + Yrs_in_NHL + Wt + GP, data = scaled_data)
 summary(baseline)
@@
 ```{r}
-plot(Yhat, e, type = 'n', main = 'Fitted vs Residual')
-grid()
-abline(h = 0, lwd = 2, col = 'gray')
-points(Yhat, e, pch = 16, col = 'darkblue')
-```
-
-```{r}
-qq_model  <- qqnorm(es, plot.it = FALSE)
-qqplot(qq_model$x, qq_model$y, xlab = 'E(e*)', ylab = 'e*', pch = 16, col = 'darkblue',
-       main = 'Q-Q Plot')
-qqline(qq_model$y, col = 'gray', lwd = 2)
-grid()
-```
-
-```{r}
 vif(baseline)
@@
 
+### Cross-Nested Validation
 ```{r}
 library(glmnet)
-X         <- model.matrix(Salary_log ~ G + A + GF + plus_mins + TOI + iSF + Yrs_in_NHL + Wt + GP -1, data = scaled_data)
+library(nestedcv)
+X         <- model.matrix(Salary_log ~ A + GF + plus_mins + TOI + iSF + Yrs_in_NHL + Wt + GP -1, data = scaled_data)
 Y         <- scaled_data$Salary_log
 
-ridge_cv   <- cv.glmnet(x = X, y = Y, family = "gaussian", alpha = 0, type.measure = "mse", nfolds = nrow(scaled_data), grouped = FALSE)
+test <- nestcv.glmnet(x = X, y = Y, family = "gaussian", alphaSet = seq(0, 0.3, 0.1), type.measure = "mse", n_outer_folds = 10, n_inner_folds = 10, grouped = FALSE)
 
-plot(ridge_cv)
+plot(test$outer_result[[1]]$cvafit)
+
+```
+```{R}
+summary(test)
 ```
 
+### Penalized Regression
+
 ```{r}
-coef(ridge_cv)
+library(psych)
+describe(data$Salary)
 ```
+```{r}
+hist(data$Salary)
+```
+
+```{r, fig.height=10}
+plot(EDA[, 1:7])
+```

### 2024-12-16 Final/paper.tex
Squashed commit of the following:
commit 68f8653b09ea6801b0fbe145e60b4e10a5c68dfa
Author: mernaghp <bigbenshuffleboard@gmail.com>
Date:   Mon Dec 16 17:22:09 2024 -0500

    Paper minor edits

commit c97b16e7507fe799428ff684a2bc597bf43c5156
Author: mernaghp <bigbenshuffleboard@gmail.com>
Date:   Mon Dec 16 15:47:04 2024 -0500

    More paper revisions

commit ebd2ccd875820546915c30fee9b15df9ca2fabeb
Merge: 95fd5b2 1136e8b
Author: mernaghp <bigbenshuffleboard@gmail.com>
Date:   Mon Dec 16 13:41:09 2024 -0500

    Merge branch 'main' into pm

commit 95fd5b2f444f19850552e204951e69ab4fd56681
Autho

\documentclass[10pt]{article}
\usepackage{sdss2020} % Uses Times Roman font (either newtx or times package)
\usepackage{url}
\usepackage{latexsym}
\usepackage{amsmath, amsthm, amsfonts}
\usepackage{algorithm, algorithmic}  
\usepackage{multirow,graphicx}
\usepackage[flushleft]{threeparttable}
\usepackage[T1]{fontenc}
\usepackage{booktabs}
\usepackage{siunitx} 
\usepackage[hidelinks]{hyperref}
\usepackage{supertabular}
\usepackage{xurl}
\usepackage{xcolor}
\usepackage{caption}
%\usepackage{multicols}

\graphicspath{ {./Images/} }
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\renewcommand{\thesubsection}{\thesection.\Alph{subsection}}
\renewcommand\fbox{\fcolorbox{gray}{white}}

\title{Modeling NHL Player Salaries: Evaluating Expectations}

\author{
  C. Logue \\
  Georgetown University \\
  Washington, DC \\
\\\And
  P. Mernagh \\
  Georgetown University \\
  Washington, DC \\
\\\And
  Z. Holden \\
  Georgetown University \\
  Washington, DC \\
\\}

\begin{document}
\maketitle

\section{Introduction}
The study aims to model NHL player salaries as a function of their key performance metrics. We are motivated by the availability of data around players’ offensive performance (e.g., shots, goals, and assists),  defensive performance (e.g., blocked shots and hits), position and team data points, and general demographic and physical characteristics. More advanced hockey statistics such as Corsi, Fenwick, shooting-plus-save percentage (PDO), and Expected Goals (xG) are also available for consideration. Since salaries are contractual and fixed before a season, we effectively evaluate players’ subsequent performance and determine whether they lived up to expectations. Though not dedicated hockey analysts, we have an initial understanding of potentially significant metrics that we wish to either confirm or invalidate. Our analysis accounts for some peculiarities of NHL finance, specifically entry-level contracts for rookies, but does not aim to capture the effects of signing bonuses, the salary cap, multi-year contracts, or players’ movements between the NHL and AHL. Nevertheless, we expect to discover key drivers of players’ salaries.

\section{Data}
\subsection{Collection}
Our data was created by Rob Vallman, a former member of the Professional Hockey Writers Association, and an early author and blogger in the Hockey Analytics space. In 2017, Vallman publicly released a dataset with 154 season-level performance metrics for 874 players (non-goalies) in the prior NHL season (2016 - 2017). Vallman’s blog has long been taken down, but we were able to collect a redistribution of the data on Kaggle, an open-source data science platform which hosts datasets and competitions.

\subsection{Cleaning}
The data upon receipt had been split into train and test sets; we begin by recombining. Player salaries are converted to millions of dollars, and the minimal number of null values are imputed with the median value of their respective columns. Outliers are identified but left alone. Players within the first three years of their NHL career are dropped, provided they are on fixed entry-level contracts that don’t appropriately reflect a club’s expectation. The players' years in the NHL (Yrs\_in\_NHL) are computed as 2016 - DftYr, their original draft year. As such, undrafted players are dropped, provided their contract form can’t be controlled for. Finally, we made the assumption that offensive players and defensive players would have fundamentally different expectations for their on-ice contribution. To hone our analysis, we decided to drop defensive players (Pos: D), instead opting to focus on offensive players (Pos: LW, RW, C). Our final cut was 382 hockey players.

\subsection{Summary Statistics}
Summary statistics for initially considered variables are found in \autoref{tab:summary_stats}. A complete data dictionary can be found in \autoref{a:data-dictionary}. 
\begin{table}[ht]
\caption{\label{tab:summary_stats}Summary Statistics (n =382)}
\centering
\begin{tabular}[t]{lrrr}
\toprule
  & mean & min & max\\
\midrule
Assists & 15.4 & 0.0 & 63.0\\
Corsi Against & 776.9 & 4.0 & 1678.0\\
Corsi For & 844.8 & 6.0 & 1983.0\\
Fenwick Against & 579.4 & 4.0 & 1264.0\\
Fenwick For & 629.8 & 5.0 & 1462.0\\
Faceoff Pct & 41.2 & 0.0 & 100.0\\
Goals & 11.5 & 0.0 & 44.0\\
Goals Against & 36.0 & 0.0 & 91.0\\
Goals For & 40.7 & 0.0 & 120.0\\
Games Played & 57.7 & 1.0 & 82.0\\
Grit & 124.0 & 0.0 & 462.0\\
Height & 72.8 & 66.0 & 78.0\\
Blocks & 26.9 & 0.0 & 99.0\\
Hits Taken & 62.1 & 0.0 & 186.0\\
Hits Dished & 67.5 & 0.0 & 300.0\\
Shots & 105.0 & 0.0 & 313.0\\
PDO & 992.2 & 500.0 & 1257.0\\
Penalties in min. & 28.5 & 0.0 & 133.0\\
Plus-Minus & -0.8 & -34.0 & 34.0\\
Salary & 2.9 & 0.6 & 14.0\\
Time on Ice & 52732.2 & 528.0 & 105263.0\\
Weight & 200.6 & 157.0 & 244.0\\
xGA & 36.9 & 0.2 & 90.5\\
xGF & 41.5 & 0.1 & 111.1\\
Years in NHL & 8.7 & 4.0 & 26.0\\
\bottomrule
\end{tabular}
\end{table}

\section{Methods}
\subsection{Benchmark}
Our first objective is to construct a simple “baseline” model that can serve as a benchmark for alternative models. Fifteen features are carefully chosen along three dimensions we believe should be well-compensated: offensive production, defensive contribution, and overall physicality. The relationships between these covariates and the outcome of interest is explored. Pair plots exhibit positive, reasonably linear relationships between salary and the offensive covariates—Goals, Assists, Shots, Goals For, Goals Against and Plus Minus. Defensive covariates—Goals Against, Blocks, and Grit—show little to no relationship; while Years in the NHL, Height, Weight, Games Played and Time on Ice sit somewhere in the middle. Preliminary model fits demonstrate clear heteroscedasticity of residuals, along with non-normality characterized by heavy tails. Log transformation is applied to correct for this.
Multiple linear regression is fit, and residual diagnostics are performed. Variance Inflation Factors are estimated to formally evaluate multicollinearity. Finally, penalized methods are explored to perform model selection, as well as resolve multicollinearity.

\subsection{Advanced Hockey Statistics}
In NHL analytics, “advanced” statistics are broadly defined as metrics that expand on the complexity of the traditional box-score statistics like goals, shots, hits, penalty minutes, or plus-minus differential. Traditional box-score statistics are typically combined with additional in-game measurements to provide a more nuanced understanding of how player performance—evaluated in terms of high-level concepts such as shot danger, quality of competition, and even “luck” (or lack thereof)—impacts key game outcomes. Their popularity has exploded in the past decade with in-house NHL franchise analytics departments to inform player contract offers, so we anticipate that their inclusion will increase the predictive power of the benchmark model. For simplicity in analysis, we choose to only consider the four fundamental advanced statistics, namely Corsi (C), Fenwick (F), expected goals (xG), and shooting-plus-save percentage (PDO).
To begin, we fit an initial saturated multiple linear regression model with the traditional box-score statistics from the benchmark model and the fundamental advanced statistics and perform residual and diagnostic analyses. Since advanced statistics are frequently considered together to provide complete context and cover blind spots from other metrics, we also incorporate interaction terms to capture established relationships between the offensive and defensive components of each advanced statistic, Corsi and Fenwick, and xG and PDO. There is significant overlap in the feature space among some of the advanced statistics—for example, Corsi and Fenwick both measure the sum of all shot attempts, except Fenwick excludes blocked shot attempts—so we apply a second round of LASSO penalized regression to perform model selection and resolve the reintroduced multicollinearity. 
Moreover, we observe noticeable heteroscedasticity in both the benchmark and advanced statistics models and therefore perform generalized least squares regression with both sets of covariates. Residual analysis demonstrates that this is likely caused by games played (\autoref{advanced-games}),
and we accordingly impose a heterogeneous variance pattern grouped by games played. The games are spaced at approximately equal intervals of time (excluding the mandatory mid-season bye week), and performance metrics are expected to serially correlate, as their increases and decreases usually happen together with slumps or streaks. Thus, we impose a first-order autoregressive, or AR(1), correlation pattern.

\subsection{GAM}
Since the benchmark model did not consider any nonlinear approaches, we finally assess whether any covariates can be appropriately modeled with a smoothed effect. We fit an initial GAM with REML and a smoother for every covariate, and perform residual and basis dimension diagnostics for each smoothed effect. We then fit a revised model, maintaining a smoothed effect for a small subset of the covariates. We also consider the case of dropping a potential outlier with respect to Years in the NHL.

\section{Results}
\subsection{Benchmark}
The initial OLS model %(\autoref{tab:initial} in \autoref{a:supplemental}) 
exhibits a strong degree of predictive power, with an adjusted coefficient of determination of 0.693. In residual analysis (\autoref{benchmark-residuals}), we observe a good degree of random scatter, with no trend in LOWESS that might suggest non-linearity in the regression function. Heteroskedasticity remains an issue, even with log transformation. There are no outliers. The QQ plot (\autoref{benchmark-qq}) exhibits normality, the result of an effective log transformation. Of the original fifteen covariates, the model shows strong evidence suggesting a linear relationship between Salary and only three—Time on Ice, Years in the NHL and Games Played. NHL career seems the most intuitive—for every year spent in the league, we observe a statistically significant increase in Salary by a factor of 1.06, or 6\%, on average, all else held equal. 
 In other words, there is a nice baseline “raise” for every year spent in the NHL. Beyond this, there are multiple signs of instability, of which many are recognized as informal signs of multicollinearity. In sensitivity analysis, many coefficients show large variation in both sign and magnitude—for example, when Goals is removed and the model re-fit, Assists flips to a positive factor, with Goals For cut in half. Strong, positive pairwise correlation is observed within the design matrix.
With a mean VIF = 14.91, there is strong evidence confirming influential collinearity in the predictors of interest; however, as we can observe in the individual VIF scores, multicollinearity is not consistently distributed among each covariates. Instead, correlation falls within stronger and weaker pockets of covariates, directly observable in the pairwise correlation plot—a stronger pocket emerges around offensive features, a slightly weaker among the defensive, and a little couplet around height and weight. Provided we have a high number of predictors, varying levels of importance and pockets of high correlation, we believe an elastic net is best-suited for penalized regression.
Elastic net is performed (\autoref{tab:benchmark}), showing a reasonably stable shrinkage. The lambda values are quite small, pointing to model selection as the best use case—we induce bias, but don’t observe any meaningful drop in the variance tradeoff. Iterating through different levels of $\alpha$, we settle on the traditional value at $\alpha = 0.5$, with the $\lambda$ chosen within the first standard error; introducing the $L_2$ norm to enforce a “Grouping effect,” while leaning more heavily towards sparsity. Altogether, the model performs well along these expectations, preserving the group of offensive covariates, and shrinking out the unimportant group of defensive features, as we had expected might be the case. 
Comparing the shrinkage effect to the original OLS model (\autoref{tab:initial}), we observe correction in sign and magnitude that is perhaps more reasonable for interpretation. Assists are strongly corrected in the positive direction, and share similar impact to goals for—for every assist registered, we observe an increase in salary by a factor of about 1.1\%. This suggests that salary may be well-aligned with offensive production. Games played shrinks much closer to a neutral effect but remains negative.

\begin{table}[ht]
\caption{\label{tab:benchmark}Benchmark Elastic Net}
\centering
\begin{threeparttable}
\begin{tabular}[t]{lc}
\toprule
 & s1 \\ 
  \midrule
(Intercept) & 0.2676 \\ 
  Goals & $\cdots$ \\ 
  Assists & 1.0107 \\ 
  Goals For & 1.0115  \\ 
  Goals Against & $\cdots$  \\ 
  Plus Minus & 0.9918 \\ 
  Time on Ice & 1.0000  \\ 
  Shots & 1.0022  \\ 
  Years in NHL & 1.0530 \\ 
  Height & $\cdots$ \\ 
  Weight & 1.0037 \\ 
  Hits Dished &  $\cdots$ \\ 
  Hits Taken &  $\cdots$ \\ 
  Blocks &  $\cdots$   \\ 
  Grit &  $\cdots$  \\ 
  Games Played & 0.9951 \\ 
\bottomrule
\end{tabular}
    \begin{tablenotes}
      \item $\alpha = 0.5, \lambda = 0.04$
    \end{tablenotes}
  \end{threeparttable}    
\end{table}

Ultimately, elastic net is effective, but doesn’t perfectly preserve groups—goals fall out of the offensive group, with height falling out of the physical characteristics grouping. To correct for this, we explored a penalized method called Group Lasso, introduced in 2006 for experiments in which a grouping of covariates is “natural.” The objective function is similar to Elastic Net in that it applies both the $L_1$ and $L_2$ norms; however, it first partitions the feature space by pre-defined “groupings,” and implements penalization separately within the subspaces. The model specification can be found in \autoref{tab:benchmark-grouped}. The model does correct the Elastic Net in preserving Goals and Height; however, the exponentiated coefficient estimates are inflated significantly. For the sake of reliability, we move forward with Elastic Net as our benchmark.

\subsection{Advanced Hockey Statistics}
The initial multiple linear regression model with advanced statistics (\autoref{tab:advanced}) demonstrates improved predictive power over the benchmark model, achieving an adjusted coefficient of determination $R^2_a$ = 0.7161. In general, the interaction terms exhibit higher significance—measured in terms of p-value—than the underlying first-order covariates, which supports the modern conception of the four fundamental advanced statistics as starting metrics to drive compound analysis with other advanced statistics rather than meaningful metrics for independent analysis. That being said, our residual and diagnostic analyses indicate that it has two primary shortcomings that need to be remediated. 
First, the reintroduced multicollinearity is substantial, as indicated by the mean VIF = 182.6531, so we select a sparse representation of the model via LASSO regression (\autoref{tab:advanced-LASSO}), opting for the $L_1$ penalty corresponding to a cross-validated error within one standard error of the minimum given by $\lambda$ = 0.0151. Similar to the benchmark model, we find that this regularization largely favors the covariates closely related to offensive performance (e.g., team expected goals while the player was on the ice (xGF), team shot attempts while the player was on the ice (CF)) while shrinking the corresponding defensive components entirely out of the model.
Second, the heteroscedasticity observed in the residuals of the benchmark model conveys to the advanced statistics model (\autoref{advanced-residuals}), though the QQ plot still exhibits normality (\autoref{advanced-qq}). Accordingly, we perform GLS regression with both models (\autoref{tab:benchmark-gls} and \autoref{tab:advanced-gls} respectively), specifying the variance pattern as heterogeneous grouped by games played and the correlation pattern as first-order autoregressive. Plotting the standardized residuals against the fitted values for log-transformed salary, we find that the non-constancy of variance has been largely resolved in each case (\autoref{baseline-gls-residuals} and \autoref{advanced-gls-residuals} respectively). We then calculate the residual standard error ($se$) and Akaike information criterion for comparison, obtaining $se$ = 0.4217 and AIC = 709.1128 with the benchmark model and $se$ = 0.3266 and AIC = 693.7423 with the advanced statistics model. Thus, we conclude with certainty that the advanced statistics GLS regression model is superior to the benchmark GLS regression model in terms of both goodness-of-fit and parsimony.

\subsection{GAM}
Initial GAM results can be found in \autoref{tab.gam}%in \autoref{a:supplemental}
. A smoothed effect is modeled for Time on Ice, Years in the NHL, and Weight. Partial effect plots (\autoref{TOI-initial} - \autoref{Wt-initial}) demonstrate relatively good fit, with partial residuals falling fairly evenly around the smoothed function. However, the smoothed effect for Weight is still almost linear. Additionally, the smoothed effect for Years in the NHL is noticeably influenced in the right tail by an outlier, specifically Jaromír Jágr, who was drafted in 1990 and (as of writing) has had the longest career in professional ice hockey history. The revised GAM therefore only smooths Time on Ice and Years in the NHL, and considers the effect of dropping Jágr. Summarized results for the revised model are found in \autoref{tab.gam.revised}, with partial effect plots for Time on ice (\autoref{TOI-revised}) and Years in the NHL (\autoref{Yrs-in-NHL-revised}).  Deviance explained marginally decreases from 74.2\% to 73.9\%, but the Years in the NHL smoother is more justifiable. Significant covariates are similar between the initial and revised GAM vs. the benchmark model, with Time on Ice, Games Played, and Years in the NHL all highly significant. Concurvity analysis for both the initial and revised GAM models suggest room for improvement, similar to the multicollinearity in the benchmark model. High concurvity is seen in the offense-related covariates, and in Games Played and Time on Ice. Also similar to the benchmark model, the slightly negative Goals and Games Played coefficients are unintuitive and worth further remediation.

\section{Discussion}
Across the models considered, Assists, Goals, Shots, Time on Ice, Years in the NHL, and Games Played are generally significant. The defensive statistics we originally assumed predictive, namely Grit, Hits, and Blocks, were dropped in penalized regression, which is intuitive since we specifically modeled offensive players. We did attempt to model defensive players separately; even then, the defensive statistics were unexpectedly not very predictive. For both offensive and defensive players, Time on Ice, Years in the NHL, and Games Played remained significant.

It was not clear cut to identify or interpret the influence of team statistics on an individual player’s earnings. Goals For is significant, but by nature highly correlated with individual Goals. Plus-minus is also significant, and much less correlated with individual Goals. However, its negative coefficient is unexpected.

Additional data around player contracts, such as the form and length of entry-level contracts, the minimum salary, the team salary cap, and the typical length of multi-year contracts, would improve our modeling. Better understanding around games played, including data for games missed due to healthy scratches or to injury, as well as for  players sent down to the AHL, would aid with explanation. A more detailed data dictionary, with precise definitions and formulas for compound measures and the advanced hockey statistics, would help with interpreting their modeled effects. Finally, if we had data spanning multiple seasons, we could capture the effects of previous seasons on the renegotiation of a player’s contract.

%\bibliographystyle{sdss2020} 
%\bibliography{bibl}

\section*{References}
\flushleft{DeMonte, D. (2020, May 6). \textit{Beginner’s Guide to Advanced Hockey Statistics}.
https://sites.northwestern.edu/nusportsanalytics/2020/05/06/\\advanced-hockey-statistics\\[1em]

\flushleft{Kaggle. (n.d.). \textit{Predict NHL Player Salaries}. 
https://www.kaggle.com/datasets/camnugent/predict-nhl-player-salaries/data\\[1em]

\flushleft{Vollman, R. (n.d.). \textit{NHL 2016-17 Player Data}.
https://web.archive.org/web/20171213112940/http://\\www.hockeyabstract.com/testimonials/nhl2016-17playerdata\\[1em]

\flushleft{Zlomislic, M. (2021, December 26). \textit{Hockey advanced analytics: What are they \& why are they important?}.
https://thehockeywriters.com/hockey-advanced-analytics-importance-metrics-definitions\\[1em]

\appendix
\clearpage

\section{Data Dictionary}
\label{a:data-dictionary}

\begin{table}[H]
\centering
%\caption*{Data Dictionary}
\label{data-dictionary}
\begin{tabular}[t]{p{0.2\columnwidth} p{0.7\columnwidth} }
\toprule
 & Meaning\\
\midrule
A & Assists, equivalent to A1 + A2\\
A1 & Primary assists\\
A2 & Secondary assists\\
CA & Shot attempts allowed (Corsi, SAT) while a player was on the ice\\
CF & The team's shot attempts (Corsi, SAT) while a player was on the ice\\
DftYr & Draft Year\\
FA & Unblocked shot attempts allowed (Fenwick, USAT) while a player was on the ice\\
FF & The team's unblocked shot attempts (Fenwick, USAT) while a player was on the ice\\
FO\_pct & Faceoff winning percentage \\
G & Goals\\
GA & Goals allowed while a player was on the ice\\
GF & The team's goals while a player was on the ice\\
GP & Games played\\
Grit & Hits, blocked shots, penalty minutes, and major penalties\\
Ht & Height, in inches\\
iBLK & Shots blocked\\
iHA & Hits taken\\
iHF & Hits thrown\\
iSF & Shots on goal\\
PIM & Penalties in minutes\\
plus\_mins & Plus/minus\\
TOI & Time on ice, in seconds\\
Wt & Weight, in pounds\\
xGA & Expected goals allowed (weighted shots) while a player was on the ice\\
xGF & The team's expected goals (weighted shots) while a player was on the ice\\
Yrs\_in\_NHL & Years in NHL (2016 - DftYr) \\
\bottomrule
\end{tabular}
\end{table}

\vfill\eject
\section{Supplemental Results}
\label{a:supplemental}

\begin{table}[tph]
\caption{Benchmark Model}
\label{tab:initial}
\centering
\begin{threeparttable}
\begin{tabular}[t]{lrrrr}
  \toprule
 & Estimate & t value & Pr($>$$|$t$|$) & 95\% CI \\ 
  \midrule
(Intercept) & 0.2862  & -1.142 & 0.254 & (0.0332, 2.4685) \\
  G & 0.9825 &  -2.004 & 0.046 & (0.9657, 0.9997) \\
  A & 0.9974 &  -0.346 & 0.729 & (0.9831, 1.0120) \\ 
  GF & 1.0145 &  2.562 & 0.011 & (1.0033, 1.0257) \\ 
  GA & 0.9939 & -1.305 & 0.193 & (0.9849, 1.0031) \\ 
  plus\_mins & 0.9907 &  -2.468 & 0.014 & (0.9833, 0.9981) \\ 
  TOI & 1.0000 &  4.812 &  $<$0.001 & (1.0000, 1.0000) \\
  iSF & 1.0024 & 2.166 & 0.031 & (1.0002, 1.0046) \\ 
  Years\_in\_ NHL & 1.0603 & 8.103 &  $<$0.001 & (1.0453, 1.0755) \\
  Ht & 0.9962 &  -0.196 & 0.844 & (0.9588, 1.0350) \\ 
  Wt & 1.0052  & 1.772 & 0.077 & (0.9994, 1.0109) \\ 
  iHF & 0.9995  & -0.278 & 0.781 & (0.9957, 1.0033) \\ 
  iHA & 1.0005  & 0.429 & 0.668 & (0.9981, 1.0030) \\ 
  iBLK & 0.9959  & -1.526 & 0.128 & (0.9908, 1.0012) \\ 
  Grit & 1.0013  & 0.916 & 0.360 & (0.9985, 1.0041) \\ 
  GP & 0.9740 & -7.210 & $<$0.001 &  (0.9670, 0.9810) \\ 
   \bottomrule
\end{tabular}
    \begin{tablenotes}
      \item $R^2_a$ = 0.69
    \end{tablenotes}
  \end{threeparttable} 
\end{table}

\begin{table}[tph]
\caption{\label{tab:benchmark-grouped}Benchmark Grouped LASSO}
\centering
\begin{threeparttable}
\begin{tabular}[t]{lc}
\toprule
 & s1 \\ 
  \midrule
(Intercept) & 1.9515\\ 
  Goals & 1.0597\\ 
  Assists & 1.2013 \\ 
  Goals For & 1.2367  \\ 
  Goals Against & $\cdots$ \\ 
  Plus Minus & 0.9236 \\ 
  Time on Ice & 1.1751 \\ 
  Shots & 1.1717  \\ 
  Years in NHL & 1.2207 \\ 
  Height & 1.0076 \\ 
  Weight & 1.0295 \\ 
  Hits Dished &  $\cdots$ \\ 
  Hits Taken &  $\cdots$ \\ 
  Blocks &  $\cdots$   \\ 
  Grit &  $\cdots$  \\ 
  Games Played & 0.8856 \\ 
\bottomrule
\end{tabular}
    \begin{tablenotes}
      \item $\lambda = 0.03$
    \end{tablenotes}
  \end{threeparttable}    
\end{table}

%\begin{table}[tph]
%\caption{\label{tab:advanced}Advanced Metrics Initial Model}
%\centering
%\begin{threeparttable}
%\begin{tabular}[t]{lrrrrr}
%  \toprule
%  & Estimate & Std. Error & t value & Pr($>$$|$t$|$) & 95\% CI \\
%      \hline
%    (Intercept) & 0.2844 & 0.0682 & 4.17 & $<$0.0001 & (0.1503, 0.4184) \\ 
%    A & 0.0903 & 0.0871 & 1.04 & 0.3004 & (-0.0809, 0.2616) \\ 
%    GF & 0.0835 & 0.2150 & 0.39 & 0.6978 & (-0.3392, 0.5063) \\ 
%    Plus-Minus & -0.1309 & 0.0495 & -2.65 & 0.0085 & (-0.2282, -0.0336) \\ 
%    TOI & 1.0859 & 0.3806 & 2.85 & 0.0046 & (0.3375, 1.8343) \\ 
%    iSF & 0.1505 & 0.0800 & 1.88 & 0.0608 & (-0.0068, 0.3077) \\ 
%    Yrs\_in\_NHL & 0.2537 & 0.0293 & 8.65 & <0.0001 & (0.1960, 0.3114) \\ 
%    Wt & 0.0927 & 0.0292 & 3.17 & 0.0016 & (0.0352, 0.1501) \\ 
%    GP & -1.2380 & 0.1479 & -8.37 & <0.0001 & (-1.5289, -0.9470) \\ 
%    PDO & 0.2387 & 0.1320 & 1.81 & 0.0714 & (-0.0209, 0.4983) \\ 
%    CA & 1.2701 & 0.5870 & 2.16 & 0.0311 & (0.1157, 2.4244) \\ 
%    CF & 0.6446 & 0.7586 & 0.85 & 0.3961 & (-0.8473, 2.1364) \\
%    FA & -1.0774 & 0.6302 & -1.71 & 0.0882 & (-2.3167, 0.1619) \\ 
%    FF & 0.2472 & 0.8590 & 0.29 & 0.7737 & (-1.4420, 1.9363) \\ 
%    xGA & -0.1335 & 0.2415 & -0.55 & 0.5807 & (-0.6084, 0.3414) \\ 
%    xGF & -0.6520 & 0.3075 & -2.12 & 0.0347 & (-1.2568, -0.0472) \\ 
%    CA:CF & -0.1448 & 0.4847 & -0.30 & 0.7653 & (-1.0980, 0.8084) \\ 
%    FA:FF & 0.6112 & 0.5099 & 1.20 & 0.2314 & (-0.3915, 1.6139) \\ 
%    CA:FA & -0.5539 & 0.1634 & -3.39 & 0.0008 & (-0.8752, -0.23262936) \\ 
%    CF:FF & -0.4170 & 0.1468 & -2.84 & 0.0047 & (-0.7056, -0.1284) \\ 
%    xGA:xGF & 0.2437 & 0.1832 & 1.33 & 0.1842 & (-0.1165, 0.6038) \\ 
%    PDO:xGA & 0.0339 & 0.1366 & 0.25 & 0.8040 & (-0.2346, 0.3025) \\ 
%    PDO:xGF & 0.1089 & 0.1537 & 0.71 & 0.4789 & (-0.1932, 0.4111) \\ 
%   \bottomrule
%\end{tabular}
%    \begin{tablenotes}
%      \item $R^2_a$ = 0.7161, $df$ = 359, $se$ = 0.5328
%    \end{tablenotes}
%  \end{threeparttable}    
%\end{table}

% Remove Std. Error
%\begin{table}[tph]
%\caption{\label{tab:advanced}Advanced Statistics Initial Model (Saturated)}
%\centering
%\begin{threeparttable}
%\begin{tabular}[t]{lrrrrr}
%  \toprule
%  & Estimate & t value & Pr($>$$|$t$|$) & 95\% CI \\
%      \midrule
%    (Intercept) & 0.2844 & 4.17 & $<$0.0001 & (0.1503, 0.4184) \\ 
%    A & 0.0903 & 1.04 & 0.3004 & (-0.0809, 0.2616) \\ 
%    GF & 0.0835  & 0.39 & 0.6978 & (-0.3392, 0.5063) \\ 
%    Plus-Minus & -0.1309 & -2.65 & 0.0085 & (-0.2282, -0.0336) \\ 
%    TOI & 1.0859 & 2.85 & 0.0046 & (0.3375, 1.8343) \\ 
%    iSF & 0.1505 & 1.88 & 0.0608 & (-0.0068, 0.3077) \\ 
%    Yrs\_in\_NHL & 0.2537  & 8.65 & <0.0001 & (0.1960, 0.3114) \\ 
%    Wt & 0.0927 & 3.17 & 0.0016 & (0.0352, 0.1501) \\ 
%    GP & -1.2380 & -8.37 & <0.0001 & (-1.5289, -0.9470) \\ 
%    PDO & 0.2387 & 1.81 & 0.0714 & (-0.0209, 0.4983) \\ 
%    CA & 1.2701 & 2.16 & 0.0311 & (0.1157, 2.4244) \\ 
%    CF & 0.6446 & 0.85 & 0.3961 & (-0.8473, 2.1364) \\
%    FA & -1.0774 & -1.71 & 0.0882 & (-2.3167, 0.1619) \\ 
%    FF & 0.2472 & 0.29 & 0.7737 & (-1.4420, 1.9363) \\ 
%    xGA & -0.1335 & -0.55 & 0.5807 & (-0.6084, 0.3414) \\ 
%    xGF & -0.6520 & -2.12 & 0.0347 & (-1.2568, -0.0472) \\ 
%    CA:CF & -0.1448 & -0.30 & 0.7653 & (-1.0980, 0.8084) \\ 
%    FA:FF & 0.6112  & 1.20 & 0.2314 & (-0.3915, 1.6139) \\ 
%    CA:FA & -0.5539 & -3.39 & 0.0008 & (-0.8752, -0.2326) \\ 
%    CF:FF & -0.4170 & -2.84 & 0.0047 & (-0.7056, -0.1284) \\ 
%    xGA:xGF & 0.2437 & 1.33 & 0.1842 & (-0.1165, 0.6038) \\ 
%    PDO:xGA & 0.0339 & 0.25 & 0.8040 & (-0.2346, 0.3025) \\ 
%    PDO:xGF & 0.1089 & 0.71 & 0.4789 & (-0.1932, 0.4111) \\ 
%   \bottomrule
%\end{tabular}
%    \begin{tablenotes}
%      \item $R^2_a$ = 0.72
%    \end{tablenotes}
%  \end{threeparttable}    
%\end{table}

% Remove CI
\begin{table}[tph]
\caption{Advanced Statisics Initial Model}
\label{tab:advanced}
\centering
\begin{threeparttable}
\begin{tabular}[t]{lrrrr}
  \toprule
  & Estimate & Std. Error & t value & Pr($>$$|$t$|$)\\
      \midrule
    (Intercept) & 0.2844 & 0.0682 & 4.17 & <0.0001 \\ 
    A & 0.0903 & 0.0871 & 1.04 & 0.3004 \\ 
    GF & 0.0835 & 0.2150 & 0.39 & 0.6978 \\ 
    Plus-Minus & -0.1309 & 0.0495 & -2.65 & 0.0085 \\ 
    TOI & 1.0859 & 0.3806 & 2.85 & 0.0046 \\ 
    iSF & 0.1505 & 0.0800 & 1.88 & 0.0608 \\ 
    Yrs\_in\_NHL & 0.2537 & 0.0293 & 8.65 & <0.0001 \\ 
    Wt & 0.0927 & 0.0292 & 3.17 & 0.0016 \\ 
    GP & -1.2380 & 0.1479 & -8.37 & <0.0001 \\ 
    PDO & 0.2387 & 0.1320 & 1.81 & 0.0714 \\ 
    CA & 1.2701 & 0.5870 & 2.16 & 0.0311 \\ 
    CF & 0.6446 & 0.7586 & 0.85 & 0.3961 \\
    FA & -1.0774 & 0.6302 & -1.71 & 0.0882 \\ 
    FF & 0.2472 & 0.8590 & 0.29 & 0.7737 \\ 
    xGA & -0.1335 & 0.2415 & -0.55 & 0.5807 \\ 
    xGF & -0.6520 & 0.3075 & -2.12 & 0.0347 \\ 
    CA:CF & -0.1448 & 0.4847 & -0.30 & 0.7653 \\ 
    FA:FF & 0.6112 & 0.5099 & 1.20 & 0.2314 \\ 
    CA:FA & -0.5539 & 0.1634 & -3.39 & 0.0008 \\ 
    CF:FF & -0.4170 & 0.1468 & -2.84 & 0.0047 \\ 
    xGA:xGF & 0.2437 & 0.1832 & 1.33 & 0.1842 \\ 
    PDO:xGA & 0.0339 & 0.1366 & 0.25 & 0.8040 \\ 
    PDO:xGF & 0.1089 & 0.1537 & 0.71 & 0.4789 \\ 
   \bottomrule
\end{tabular}
    \begin{tablenotes}
      \item $R^2_a$ = 0.72  % 0.7161, $df$ = 359, $se$ = 0.5328
    \end{tablenotes}
  \end{threeparttable}    
\end{table}

\begin{table}[tph]
\caption{\label{tab:advanced-LASSO}Advanced Statistics LASSO Model}
\centering
\begin{threeparttable}
   \begin{tabular}[t]{lr}
      \toprule
     & s1 \\
      \midrule
    (Intercept) & 0.0836 \\
    A & 0.0945 \\
    GF & 0.2345 \\
    plus\_mins & -0.0679 \\
    TOI & 0.1861 \\
    iSF & 0.1132 \\
    Yrs\_in\_NHL & 0.2367 \\
    Wt & 0.0678 \\
    GP & -0.4374 \\
    PDO & 0.0087 \\
    CA & $\cdots$ \\
    CF & 0.4549 \\
    FA & $\cdots$ \\
    FF & $\cdots$ \\
    xGA & $\cdots$ \\
    xGF & $\cdots$ \\
    CA:CF & $\cdots$ \\
    FA:FF & $\cdots$ \\
    CA:FA & -0.0309 \\
    CF:FF & -0.0530 \\
    xGA:xGF & $\cdots$ \\
    PDO:xGA & $\cdots$ \\
    PDO:xGF  & $\cdots$ \\
    \bottomrule
    \end{tabular}
    \begin{tablenotes}
      \item  $\alpha$ = 1, $\lambda$ = 0.0151
    \end{tablenotes}
  \end{threeparttable}    
\end{table}

\begin{table}[tph]
\caption{\label{tab:benchmark-gls}Benchmark GLS Model}
\centering
\begin{threeparttable}
\begin{tabular}[t]{lrrrr}
  \toprule
     & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \midrule
    (Intercept) & -0.0620 & 0.0188 & -3.2835 & 0.0011 \\
    A & 0.0603 & 0.0589 & 1.0236 & 0.3067 \\
    GF & 0.3003 & 0.1033 & 2.9077 & 0.0039 \\
    Plus-Minus & -0.1223 & 0.0275 & -4.4414 & <0.0001 \\
    TOI & 0.9958 & 0.1175 & 8.4716 & $<$0.0001 \\
    iSF & 0.1042 & 0.0596 & 1.7469 & 0.0815 \\
    Yrs\_in\_NHL & 0.2384 & 0.0198 & 12.0294 & <0.0001\\
    Wt & 0.0570 & 0.0120 & 4.7497 & <0.0001 \\
    GP & -0.6863 & 0.0590 & -11.6330 & <0.0001\\
   \bottomrule
\end{tabular}
    \begin{tablenotes}
      \item AIC = 709.11
    \end{tablenotes}
  \end{threeparttable}    
\end{table}

\begin{table}[tph]
\caption{\label{tab:advanced-gls}Advanced Statistics GLS Model}
\centering
\begin{threeparttable}
\begin{tabular}[t]{lrrrr}
  \toprule
     & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \midrule
    (Intercept) & 0.3141 & 0.0219 & 14.3129 & <0.0001\\
    A & 0.0488 & 0.0594 & 0.8226 & 0.4112 \\
    GF & 0.0817 & 0.1131 & 0.7224 & 0.4705 \\
    plus\_mins & -0.0552 & 0.0288 & -1.9138 & 0.0564 \\
    TOI & 1.1786 & 0.2429 & 4.8521 & $<$0.0001 \\
    iSF & 0.1678 & 0.0610 & 2.7482 & 0.0063 \\
    Yrs\_in\_NHL & 0.1959 & 0.0165 & 11.8570 & $<$0.0001 \\
    Wt & 0.0513 & 0.0142 & 3.6009 & 0.0004 \\
    GP & -1.3975 & 0.0585 & -23.8883 & $<$0.0001 \\
    PDO & 0.0007 & 0.0178 & 0.0421 & 0.9664 \\
    CA & 0.4754 & 0.3046 & 1.5607 & 0.1195 \\
    CF & 1.2594 & 0.4610 & 2.7318 & 0.0066 \\
    FA & -0.1993 & 0.3269 & -0.6097 & 0.5424 \\
    FF & -1.0619 & 0.4907 & -2.1636 & 0.0311 \\
    CA:FA & -0.3578 & 0.0304 & -11.7530 & $<$0.0001 \\
    CF:FF & 0.0136 & 0.0305 & 0.4460 & 0.6558 \\
   \bottomrule
\end{tabular}
    \begin{tablenotes}
      \item AIC = 693.74
    \end{tablenotes}
  \end{threeparttable}    
\end{table}

\begin{table}[tph]
\centering
\caption{Initial GAM} 
\label{tab.gam}
\begin{threeparttable}
\begin{tabular}[t]{L{0.27\linewidth}rrrr} 
   \toprule
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\ 
   \midrule
  (Intercept) & 0.6686 & 0.0246 & 27.2225 & $<$ 0.0001 \\ 
   \midrule
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
   \midrule
  s(G) & 1.0001 & 1.0002 & 4.8506 & 0.0283 \\ 
  s(A) & 1.0000 & 1.0001 & 0.0048 & 0.9450 \\ 
  s(GF) & 1.0008 & 1.0014 & 6.2768 & 0.0126 \\ 
  s(plus\_mins) & 1.1457 & 1.2784 & 2.9811 & 0.0569 \\ 
  s(TOI) & 3.8517 & 4.8294 & 11.2133 & $<$ 0.0001 \\ 
  s(iSF) & 1.0000 & 1.0000 & 9.9730 & 0.0017 \\ 
  s(Yrs\_in\_NHL) & 3.2940 & 4.1439 & 25.0174 & $<$ 0.0001 \\ 
  s(Wt) & 2.0294 & 2.6114 & 5.0137 & 0.0031 \\ 
  s(GP) & 1.0000 & 1.0001 & 77.7958 & $<$ 0.0001 \\ 
   \bottomrule
\end{tabular}
    \begin{tablenotes}
      \item Deviance explained = 74.2\%
    \end{tablenotes}
  \end{threeparttable}    
\end{table}

\begin{table}[tph]
\centering
\caption{Revised GAM} 
\label{tab.gam.revised}
\begin{threeparttable}
%\scriptsize
\begin{tabular}[t]{L{0.27\linewidth}rrrr} 
   \toprule
A. parametric coefficients & Estimate & Std. Error & t-value & p-value \\  
  \midrule
  (Intercept) & 0.9463 & 0.4591 & 2.0612 & 0.0400 \\ 
  G & -0.0179 & 0.0081 & -2.2001 & 0.0284 \\ 
  A & 0.0008 & 0.0069 & 0.1100 & 0.9125 \\ 
  GF & 0.0130 & 0.0052 & 2.5194 & 0.0122 \\ 
  plus\_mins & -0.0061 & 0.0030 & -2.0545 & 0.0406 \\ 
  iSF & 0.0031 & 0.0010 & 3.1489 & 0.0018 \\ 
  Wt & 0.0066 & 0.0019 & 3.4912 & 0.0005 \\ 
  GP & -0.0393 & 0.0045 & -8.7432 & $<$ 0.0001 \\ 
   \midrule
B. smooth terms & edf & Ref.df & F-value & p-value \\ 
  \midrule
  s(TOI) & 3.8041 & 4.7758 & 11.0928 & $<$ 0.0001 \\ 
  s(Yrs\_in\_NHL) & 3.0052 & 3.7576 & 27.4259 & $<$ 0.0001 \\ 
   \bottomrule
\end{tabular}
    \begin{tablenotes}
     %\scriptsize
      \item Deviance explained = 73.9\%
    \end{tablenotes}
  \end{threeparttable}    
\end{table}

%\begin{table}
%\centering
%\caption{Basis Dimension Check, Initial GAM} 
%\label{tab.basis.check.initial}
%\begin{tabular}[t]{lrrrr}
%\hline
%& k\textquotesingle & edf & k-index & p-value\\
%\hline
%s(G) & 9.00 & 1.00 & 1.00 & 0.490\\
%s(A) & 9.00 & 1.00 & 1.09 & 0.965\\
%s(GF) & 9.00 & 1.00 & 0.97 & 0.260\\
%s(plus\_mins) & 9.00 & 1.15 & 0.93 & 0.045\\
%s(TOI) & 9.00 & 3.85 & 0.99 &  0.420\\
%s(iSF) & 9.00 & 1.00 & 1.02 & 0.615\\
%s(Yrs\_in\_NHL) & 9.00 & 3.29 & 0.98 & 0.310\\
%s(Wt) & 9.00 & 2.03 & 1.04 & 0.760\\
%s(GP) & 9.00 & 1.00 & 0.96 & 0.195\\
%\hline
%\end{tabular}
%\end{table}

%\begin{table}
%\centering
%\caption{Basis Dimension Check, Revised GAM} 
%\label{tab.basis.check.revised}
%\begin{tabular}[t]{lrrrr}
%\hline
%& k\textquotesingle & edf & k-index & p-value\\
%\hline
%s(TOI) & 9.00 & 3.80 & 0.99 &  0.420\\
%s(Yrs\_in\_NHL) & 9.00 & 3.01 & 0.97 & 0.280\\
%\hline
%\end{tabular}
%\end{table}
\vfill\break
\section{Supplemental Figures}
\label{a:supplemental-fig}

\begin{figure}[tph]
\centering
	\includegraphics[width=0.9\columnwidth]{games_played_vs_resid_sq}
	\caption[]
    	{\tabular[t]{@{}l@{}}Standardized Residuals vs Games Played\endtabular}
	\label{advanced-games}
\end{figure}

\begin{figure}[tph]
\centering
	\includegraphics[width=0.9\columnwidth]{benchmark_residuals_vs_fitted}
	\caption[]
    	{\tabular[t]{@{}l@{}}Residuals vs. Fitted Plot (Benchmark)\endtabular}
	\label{benchmark-residuals}
\end{figure}

\begin{figure}[tph]
\centering
	\includegraphics[width=0.9\columnwidth]{benchmark_qq}
	\caption[]
    	{\tabular[t]{@{}l@{}}Q-Q Plot (Benchmark)\endtabular}
	\label{benchmark-qq}
\end{figure}

\begin{figure}[tph]
\centering
	\includegraphics[width=0.9\columnwidth]{model_ols_adv_resid_vs_fitted_sq_adj}
	\caption[]
    	{\tabular[t]{@{}l@{}}Residuals vs. Fitted Plot \\(Advanced Statistics)\endtabular}
	\label{advanced-residuals}
\end{figure}

\begin{figure}[tph]
\centering
	\includegraphics[width=0.9\columnwidth]{model_ols_adv_qq_resid_sq_adj}
	\caption[]
    	{\tabular[t]{@{}l@{}}Q-Q Plot (Advanced Statistics)\endtabular}
	\label{advanced-qq}
\end{figure}

\begin{figure}[tph]
\centering
	\includegraphics[width=0.9\columnwidth]{model_gls_base_resid_vs_fitted_sq}
	\caption[]
    	{\tabular[t]{@{}l@{}}Residuals vs. Fitted Plot\\(Baseline GLS)\endtabular}
	\label{baseline-gls-residuals}
\end{figure}

\begin{figure}[tph]
\centering
	\includegraphics[width=0.9\columnwidth]{model_gls_adv_resid_vs_fitted_sq}
	\caption[]
    	{\tabular[t]{@{}l@{}}Residuals vs. Fitted Plot \\(Advanced Statistics GLS)\endtabular}
	\label{advanced-gls-residuals}
\end{figure}

\begin{figure}[tph]
\centering
	\fbox{\includegraphics[scale=0.4]{s(TOI)}}
	\caption[]
    	{\tabular[t]{@{}l@{}}Partial Effects Plot, Time on Ice \\ (Initial GAM)\endtabular}
	\label{TOI-initial}
\end{figure}

\begin{figure}[tph]
\centering
	\fbox{\includegraphics[scale=0.4]{s(Yrs_in_NHL)}}
	\caption[]
    	{\tabular[t]{@{}l@{}}Partial Effects Plot, Years in NHL \\ (Initial GAM)\endtabular}
	\label{Yrs-in-NHL-initial}
\end{figure}

\begin{figure}[tph]
\centering
	\fbox{\includegraphics[scale=0.4]{s(Wt)}}
	\caption[]
    	{\tabular[t]{@{}l@{}}Partial Effects Plot, Weight \\ (Initial GAM)\endtabular}
	\label{Wt-initial}
\end{figure}

\begin{figure}[tph]
\centering
	\fbox{\includegraphics[scale=0.4]{s(TOI)_revised}}
	\caption[]
    	{\tabular[t]{@{}l@{}}Partial Effects Plot, Time on Ice \\ (Revised GAM)\endtabular}
	\label{TOI-revised}
\end{figure}

\begin{figure}[tph]
\centering
	\fbox{\includegraphics[scale=0.4]{s(Yrs_in_NHL)_revised}}
	\caption[]
    	{\tabular[t]{@{}l@{}}Partial Effects Plot, Years in NHL\\ (Revised GAM)\endtabular}
	\label{Yrs-in-NHL-revised}
\end{figure}

%\clearpage
\vfill\eject
\section{Replication Data and Code}
\label{a:code}
Replication data and code can be found in the GitHub repo \mbox{https://github.com/cons-code/MATH-5500}.

\end{document}

### 2024-12-16 Final/EDA.py

@@ -1,28 +1,0 @@
-# %%
-import pandas as pd
-import seaborn as sns
-
-# %%
-df = pd.read_csv('Data/combined.csv', encoding='latin-1')
-
-sns.pairplot(df[['Salary', 'GP', 'PTS', '+/-', 
-                 'Grit', 'TOI', 'PIM', 'Is_Offense']])
-
-# %%
-numeric = df.select_dtypes(include='number')
-cols = numeric.columns.drop('Salary')
-
-for col in cols:
-    print(col, df['Salary'].corr(df[col]))
-
-# %%
-cols = numeric.columns.drop('PTS')
-for col in cols:
-    print(col, df['PTS'].corr(df[col]))
-
-# %%
-cols = numeric.columns.drop('Salary')
-msk = numeric['Salary'].ge(1)
-
-for col in cols:
-    print(col, df['Salary'][msk].corr(df[col][msk]))

### 2024-12-16 Final/R/generalized.R

@@ -1,18 +1,0 @@
-library(car)
-library(dplyr)
-library(GGally)
-library(ggplot2)
-library(ggfortify)
-
-df <- read.csv('../Data/combined.csv')
-
-model <- glm(Salary ~ TOI/GP + Is_Offense + iHF + S.Slap + iPENT + CF, 
-             data = df, family = Gamma(link = "log"))
-autoplot(model)
-coef(model)
-
-df['High.Salary'] = 1*(df['Salary'] >= 2)
-model.ge.2 <- glm(High.Salary ~ PTS + TOI/GP + Is_Offense + iHF + S.Slap + iPENT + CF,
-                  data = df, family = binomial(link = "probit"))
-coef(model.ge.2)
-autoplot(model.ge.2)

### 2024-12-16 Final/R/initial_model.R

@@ -1,10 +1,0 @@
-library(car)
-library(dplyr)
-library(GGally)
-library(ggplot2)
-library(ggfortify)
-
-df <- read.csv('../Data/combined.csv')
-model <- lm('Salary ~ GP + PTS + X... + Grit + TOI', data = df)
-summary(model)
-autoplot(model)

### 2024-12-16 Final/R/penalized.R

@@ -1,22 +1,0 @@
-library(car)
-library(dplyr)
-library(GGally)
-library(ggplot2)
-library(ggfortify)
-library(glmnet)
-
-df <- read.csv('../Data/combined.csv')
-
-# LASSO
-exclude <- c('Ovrl', 'DftRd', 'DftYr')
-clean <- df[, !names(df) %in% exclude]
-clean <- na.omit(clean)
-clean <- clean %>% dplyr::select(where(is.numeric))
-
-X <- clean[, !names(clean) %in% c('Salary')]
-Y <- log(clean$Salary)
-
-# LASSO
-model_lasso_cv  <- cv.glmnet(x = X, y = Y, family = "gaussian", alpha = 1,
-                             type.measure = "mse", nfolds = 5,
-                             grouped = FALSE)

### 2024-12-16 Final/R/weighted_least_squares.R

@@ -1,37 +1,0 @@
-library(car)
-library(GGally)
-library(ggplot2)
-
-df <- read.csv('../Data/combined.csv')
-
-# +/- is rendered as X...
-# E+/- is rendered as E...
-model_R1_log <- lm(log(Salary) ~ GP + PTS + X... + Grit + PIM, data = df)
-summary(model_R1_log)
-autoplot(model_R1_log)
-
-e_i_R1    <- resid(model_R1_log)
-y_hat_R1  <- predict(model_R1_log)
-k_hat     <- summary(model_R1_log)$sigma^2
-
-plot(y_hat_R1, abs(rstudent(model_R1_log)))
-plot(df$GP, abs(rstudent(model_R1_log)))
-plot(df$PTS, abs(rstudent(model_R1_log)))
-plot(df$X..., abs(rstudent(model_R1_log)))
-plot(df$Grit, abs(rstudent(model_R1_log)))
-plot(df$PIM, abs(rstudent(model_R1_log)))
-
-# Try with Games Played (GP)
-##### standard deviation function #####
-sd_fun_data   <- data.frame(y = abs(e_i_R1), x = df$GP)
-model_R1_sdf  <- lm(y ~ x, data = sd_fun_data)
-sigma_hat     <- predict(model_R1_sdf)
-w_i_sd        <- 1/(sigma_hat^2)
-
-model_R1_sd_wls <- lm(log(Salary) ~ GP + PTS + X... + Grit  + PIM, 
-                      data = df, weights = w_i_sd)
-round(coef(model_R1_log), 5)
-round(coef(model_R1_sd_wls), 5)
-autoplot(model_R1_sd_wls)
-mean(vif(model_R1_sd_wls))
-summary(model_R1_sd_wls)

### 2024-12-16 Final/gam_hockey.R

@@ -1,0 +1,121 @@
+library(car)
+library(dplyr)
+library(glmnet)
+library(ggfortify)
+library(ggplot2)
+library(gratia)
+library(itsadug)
+library(kableExtra)
+library(knitr)
+library(mgcv)
+library(patchwork)
+
+raw_data <- read.csv("combined.csv", header = TRUE)
+
+raw_data$Yrs_in_NHL <- 2016 - raw_data$DftYr
+raw_data$A <- raw_data$A1 + raw_data$A2
+raw_data$Salary_log <- log(raw_data$Salary)
+
+# handle missing data points
+raw_data$GF[is.na(raw_data$GF)] <- median(raw_data$GF, na.rm=TRUE)
+raw_data$GA[is.na(raw_data$GA)] <- median(raw_data$GA, na.rm=TRUE)
+raw_data$iSF[is.na(raw_data$iSF)] <- median(raw_data$iSF, na.rm=TRUE)
+raw_data$iHA[is.na(raw_data$iHA)] <- median(raw_data$iHA, na.rm=TRUE)
+raw_data$Yrs_in_NHL[is.na(raw_data$Yrs_in_NHL)] <- 0
+
+# filter out rookie players
+# only retain offense
+data <- raw_data[raw_data$Yrs_in_NHL > 3 & raw_data$Is_Offense == 1, ]
+
+# Initial GAM model
+model_initial <- gam(Salary_log ~ s(G) + s(A) + s(GF) + s(plus_mins) + s(TOI) 
+                 + s(iSF) + s(Yrs_in_NHL) + s(Wt) + s(GP), 
+                 data = data, method = "REML")
+summary(model_initial)
+
+# Output to Latex
+gamtabs(model_initial, digits = 0)
+
+# Partial Effects Plots
+p1 <- draw(model_initial, residuals = T, select = smooths(model_initial)[5]) & theme_minimal() 
+p2 <- draw(model_initial, residuals = T, select = smooths(model_initial)[7]) & theme_minimal()
+p3 <- draw(model_initial, residuals = T, select = smooths(model_initial)[8]) & theme_minimal()
+list(p1, p2, p3) |>
+  wrap_plots() +
+  plot_layout(nrow = 1, ncol = 3) 
+
+p1 & theme(text = element_text(size = 20))
+p2 & theme(text = element_text(size = 20))
+p3 & theme(text = element_text(size = 20))
+
+# Model diagnostics
[…]

### 2024-12-16 Final/glm_hockey.R

@@ -1,0 +1,35 @@
+library(boot)
+library(car)
+library(dplyr)
+library(glmnet)
+library(ggfortify)
+library(ggplot2)
+library(kableExtra)
+library(knitr)
+
+raw_data <- read.csv("combined.csv", header = TRUE)
+raw_data$Yrs_in_NHL <- 2016 - raw_data$DftYr
+
+# filter out rookie players
+# only retain offense
+data <- raw_data[raw_data$Yrs_in_NHL > 3 & raw_data$Is_Offense == 1, ]
+
+# Gamma GLM
+model <- glm(Salary ~ G + A + GF + plus_mins + TOI + iSF + 
+              Yrs_in_NHL + Wt + GP, data = data, family = Gamma(link = "log"))
+
+summary(model)
+model.diag <- glm.diag(model)
+glm.diag.plots(model, model.diag)
+
+vif(model)
+mean(vif(model))
+
+# Inverse Gaussian
+# no-go
+gaussian_inv <- glm(Salary ~ G + A + GF + plus_mins + TOI + iSF + 
+                      Yrs_in_NHL + Wt + GP, data = data, 
+                    family = gaussian(link = "inverse"))
+summary(gaussian_inv)
+gaussian_inv.diag <- glm.diag(gaussian_inv)
+glm.diag.plots(gaussian_inv, gaussian_inv.diag)

### 2024-12-16 Final/initial.py

@@ -1,400 +1,0 @@
-# %%
-import matplotlib.pyplot as plt
-import numpy as np
-import pandas as pd
-import seaborn as sns
-from sklearn.preprocessing import StandardScaler
-import statsmodels.api as sm
-
-from src.nhl_linreg import nhl_linreg
-
-# %%
-df = pd.read_csv('Data/combined.csv', encoding='latin-1')
-sns.pairplot(df[['Salary', 'GP', 'PTS', '+/-', 
-                 'Grit', 'TOI', 'PIM', 'Is_Offense']])
-plt.show();
-# %%
-# R^2_a of 0.52
-msk = df['Position'].str.contains('D')
-Y = df['Salary'][msk]
-cols = ['GP', '+/-', 'Grit', 'TOI']
-
-X = df[cols][msk]
-X = sm.add_constant(X)
-model1 = sm.OLS(Y, X)
-results1 = model1.fit()
-print(results1.summary())
-print(results1.rsquared_adj)
-diag1 = nhl_linreg(results1)
-
-vif1, fig1, ax1 = diag1()
-print(vif1['VIF Factor'].mean())
-# %%
-# R^2_a of 0.51
-cols = ['GP', 'PTS', '+/-', 'Grit', 'TOI', 'PIM'] #, 'iHF' , 'iRB', 'iSF', 'PIM']
-Y = df['Salary']
-X = df[cols]
-X = sm.add_constant(X)
-
-model2 = sm.OLS(np.log(Y), X)
-results2 = model2.fit()
-print(results2.summary())
-print(results2.rsquared_adj)
-diag2 = nhl_linreg(results2)
-
-vif2, fig2, ax2 = diag2()
-print(vif2['VIF Factor'].mean())
-
-# %%
-Y = df['Salary']
-cols = ['GP', 'PTS', '+/-', 'Grit', 'TOI', 'PIM', 'Is_Offense']
-X = df[cols]
-X = sm.add_constant(X)
-
-model3 = sm.OLS(Y, X)
-results3 = model3.fit()
-print(results3.summary())
-print(results3.rsquared_adj)
-diag3 = nhl_linreg(results3)
-vif3, fig3, ax3 = diag3()
-print(vif3['VIF Factor'].mean()) # 5.8
-
-# %%
-msk = df['Is_Offense'].eq(1)
-Y = df['Salary'][msk]
-cols = ['GP', 'PTS', '+/-', 'Grit', 'TOI', 'PIM']
[…]

### 2024-12-16 Final/src/nhl_linreg.py

@@ -1,340 +1,0 @@
-import numpy as np
-import pandas as pd
-import seaborn as sns
-import matplotlib.pyplot as plt
-
-import statsmodels
-
-from statsmodels.tools.tools import maybe_unwrap_results
-from statsmodels.graphics.gofplots import ProbPlot
-from statsmodels.stats.outliers_influence import variance_inflation_factor
-
-from typing import Type
-
-style_talk = 'seaborn-talk'    #refer to plt.style.available
-
-
-#%% NHL linear regression class & asociated utilities
-class nhl_linreg():
-
-    """
-    Diagnostic plots to identify potential problems in a linear regression fit.
-    Mainly,
-        a. non-linearity of data
-        b. Correlation of error terms
-        c. non-constant variance
-        d. outliers
-        e. high-leverage points
-        f. collinearity
-
-    Authors:
-        Prajwal Kafle (p33ajkafle@gmail.com, where 3 = r)
-        Does not come with any sort of warranty.
-        Please test the code one your end before using.
-
-        Matt Spinelli (m3spinelli@gmail.com, where 3 = r)
-        (1) Fixed incorrect annotation of the top most extreme residuals in
-            the Residuals vs Fitted and, especially, the Normal Q-Q plots.
-        (2) Changed Residuals vs Leverage plot to match closer the y-axis
-            range shown in the equivalent plot in the R package ggfortify.
-        (3) Added horizontal line at y=0 in Residuals vs Leverage plot to
-            match the plots in R package ggfortify and base R.
-        (4) Added option for placing a vertical guideline on the Residuals
-            vs Leverage plot using the rule of thumb of h = 2p/n to denote
-            high leverage (high_leverage_threshold=True).
-        (5) Added two more ways to compute the Cook's Distance (D) threshold:
-            * 'baseR': D > 1 and D > 0.5 (default)
-            * 'convention': D > 4/n
-            * 'dof': D > 4 / (n - k - 1)
-        (6) Fixed class name to conform to Pascal casing convention
-        (7) Fixed Residuals vs Leverage legend to work with loc='best'
-    """
-
-    def __init__(self,
-                 results: Type[statsmodels.regression.linear_model.RegressionResultsWrapper]) -> None:
-        """
-        For a linear regression model, generates following diagnostic plots:
-
-        a. residual
-        b. qq
[…]

### 2024-12-16 Final/weighted_least_squares.py

@@ -1,102 +1,0 @@
-# %%
-import matplotlib.pyplot as plt
-import numpy as np
-import pandas as pd
-import statsmodels.api as sm
-from statsmodels.stats.outliers_influence import OLSInfluence
-
-from src.nhl_linreg import nhl_linreg
-# %%
-df = pd.read_csv('Data/combined.csv', encoding='latin-1')
-
-# %%
-Y = df['Salary']
-cols = ['GP', 'PTS', '+/-', 'Grit', 'TOI', 'PIM', 'Is_Offense']
-X = df[cols]
-X = sm.add_constant(X)
-
-model3 = sm.OLS(Y, X)
-results3 = model3.fit()
-print(results3.summary())
-print(results3.rsquared_adj)
-diag3 = nhl_linreg(results3)
-vif3, fig3, ax3 = diag3()
-print(vif3['VIF Factor'].mean()) # 7.22
-
-# %%
-e_i = results3.resid
-y_hat = results3.fittedvalues
-k_hat = results3.scale
-
-influence = OLSInfluence(results3)
-rstudent = influence.get_resid_studentized_external()
-
-# %%
-fig, ax = plt.subplots()
-ax.scatter(y_hat, np.abs(rstudent))
-ax.set_title('Y-hat vs. |studentized residuals|')
-plt.show();
-
-# %%
-fig, ax = plt.subplots()
-ax.scatter(df['GP'], np.abs(rstudent))
-ax.set_title('Games Played vs. |studentized residuals|')
-plt.show();
-
-# %%
-fig, ax = plt.subplots()
-ax.scatter(df['PTS'], np.abs(rstudent))
-ax.set_title('Points vs. |studentized residuals|')
-plt.show();
-
-# %%
-fig, ax = plt.subplots()
-ax.scatter(df['+/-'], np.abs(rstudent))
-ax.set_title('+/- vs. |studentized residuals|')
-plt.show();
-
-# %%
-fig, ax = plt.subplots()
-ax.scatter(df['Grit'], np.abs(rstudent))
-ax.set_title('Grit vs. |studentized residuals|')
-plt.show();
-
-# %%
-fig, ax = plt.subplots()
-ax.scatter(df['TOI'], np.abs(rstudent))
-ax.set_title('Time on ice vs. |studentized residuals|')
-plt.show();
-
-# %%
-fig, ax = plt.subplots()
[…]

### 2024-12-17 Final/paper.tex
Minor cleanup

@@
 \usepackage{caption}
-%\usepackage{multicols}
 
@@
 Fenwick For & 629.8 & 5.0 & 1462.0\\
-Faceoff Pct & 41.2 & 0.0 & 100.0\\
+Faceoff Win Pct. & 41.2 & 0.0 & 100.0\\
 Goals & 11.5 & 0.0 & 44.0\\
@@
 PDO & 992.2 & 500.0 & 1257.0\\
-Penalties in min. & 28.5 & 0.0 & 133.0\\
+Penalties in Min. & 28.5 & 0.0 & 133.0\\
 Plus-Minus & -0.8 & -34.0 & 34.0\\
@@
 Weight & 200.6 & 157.0 & 244.0\\
-xGA & 36.9 & 0.2 & 90.5\\
-xGF & 41.5 & 0.1 & 111.1\\
+xG Against & 36.9 & 0.2 & 90.5\\
+xG For & 41.5 & 0.1 & 111.1\\
 Years in NHL & 8.7 & 4.0 & 26.0\\
@@
 The relationships between these covariates and the outcome of interest is explored.
-Pair plots exhibit positive, reasonably linear relationships between salary and the offensive covariates—Goals, Assists, Shots, Goals For, Goals Against and Plus Minus.
-Defensive covariates—Goals Against, Blocks, and Grit—show little to no relationship; while Years in the NHL, Height, Weight, Games Played and Time on Ice sit somewhere in the middle.
+Pair plots exhibit positive, reasonably linear relationships between salary and the offensive covariates—Goals, Assists, Shots, Goals For, and Plus-Minus.
+Defensive covariates—Goals Against, Blocks, Hits, and Grit—show little to no relationship; while Years in the NHL, Height, Weight, Games Played and Time on Ice sit somewhere in the middle.
 Preliminary model fits demonstrate clear heteroscedasticity of residuals, along with non-normality characterized by heavy tails.
@@
 In residual analysis (\autoref{benchmark-residuals}), we observe a good degree of random scatter, with no trend in LOWESS that might suggest non-linearity in the regression function.
-Heteroskedasticity remains an issue, even with log transformation.
+Heteroscedasticity remains an issue, even with log transformation.
 There are no outliers.
@@
 The lambda values are quite small, pointing to model selection as the best use case—we induce bias, but don’t observe any meaningful drop in the variance tradeoff.
-Iterating through different levels of $\alpha$, we settle on the traditional value at $\alpha = 0.5$, with the $\lambda$ chosen within the first standard error; introducing the $L_2$ norm to enforce a “Grouping effect,” while leaning more heavily towards sparsity.
+Iterating through different levels of $\alpha$, we settle on the traditional value at $\alpha = 0.5$, with the $\lambda$ chosen within the first standard error; introducing the $L_2$ norm to enforce a “grouping effect,” while leaning more heavily towards sparsity.
 Altogether, the model performs well along these expectations, preserving the group of offensive covariates, and shrinking out the unimportant group of defensive features, as we had expected might be the case.
@@
 
-%\begin{table}[tph]
-%\caption{\label{tab:advanced}Advanced Metrics Initial Model}
-%\centering
-%\begin{threeparttable}
-%\begin{tabular}[t]{lrrrrr}
-%  \toprule
-%  & Estimate & Std. Error & t value & Pr($>$$|$t$|$) & 95\% CI \\
-%      \hline
-%    (Intercept) & 0.2844 & 0.0682 & 4.17 & $<$0.0001 & (0.1503, 0.4184) \\ 
-%    A & 0.0903 & 0.0871 & 1.04 & 0.3004 & (-0.0809, 0.2616) \\ 
-%    GF & 0.0835 & 0.2150 & 0.39 & 0.6978 & (-0.3392, 0.5063) \\ 
-%    Plus-Minus & -0.1309 & 0.0495 & -2.65 & 0.0085 & (-0.2282, -0.0336) \\ 
-%    TOI & 1.0859 & 0.3806 & 2.85 & 0.0046 & (0.3375, 1.8343) \\ 
-%    iSF & 0.1505 & 0.0800 & 1.88 & 0.0608 & (-0.0068, 0.3077) \\ 
-%    Yrs\_in\_NHL & 0.2537 & 0.0293 & 8.65 & <0.0001 & (0.1960, 0.3114) \\ 
-%    Wt & 0.0927 & 0.0292 & 3.17 & 0.0016 & (0.0352, 0.1501) \\ 
-%    GP & -1.2380 & 0.1479 & -8.37 & <0.0001 & (-1.5289, -0.9470) \\ 
-%    PDO & 0.2387 & 0.1320 & 1.81 & 0.0714 & (-0.0209, 0.4983) \\ 
-%    CA & 1.2701 & 0.5870 & 2.16 & 0.0311 & (0.1157, 2.4244) \\ 
-%    CF & 0.6446 & 0.7586 & 0.85 & 0.3961 & (-0.8473, 2.1364) \\
-%    FA & -1.0774 & 0.6302 & -1.71 & 0.0882 & (-2.3167, 0.1619) \\ 
-%    FF & 0.2472 & 0.8590 & 0.29 & 0.7737 & (-1.4420, 1.9363) \\ 
-%    xGA & -0.1335 & 0.2415 & -0.55 & 0.5807 & (-0.6084, 0.3414) \\ 
-%    xGF & -0.6520 & 0.3075 & -2.12 & 0.0347 & (-1.2568, -0.0472) \\ 
-%    CA:CF & -0.1448 & 0.4847 & -0.30 & 0.7653 & (-1.0980, 0.8084) \\ 
-%    FA:FF & 0.6112 & 0.5099 & 1.20 & 0.2314 & (-0.3915, 1.6139) \\ 
-%    CA:FA & -0.5539 & 0.1634 & -3.39 & 0.0008 & (-0.8752, -0.23262936) \\ 
-%    CF:FF & -0.4170 & 0.1468 & -2.84 & 0.0047 & (-0.7056, -0.1284) \\ 
-%    xGA:xGF & 0.2437 & 0.1832 & 1.33 & 0.1842 & (-0.1165, 0.6038) \\ 
-%    PDO:xGA & 0.0339 & 0.1366 & 0.25 & 0.8040 & (-0.2346, 0.3025) \\ 
-%    PDO:xGF & 0.1089 & 0.1537 & 0.71 & 0.4789 & (-0.1932, 0.4111) \\ 
-%   \bottomrule
-%\end{tabular}
-%    \begin{tablenotes}
-%      \item $R^2_a$ = 0.7161, $df$ = 359, $se$ = 0.5328
-%    \end{tablenotes}
-%  \end{threeparttable}    
-%\end{table}
-
-% Remove Std. Error
-%\begin{table}[tph]
-%\caption{\label{tab:advanced}Advanced Statistics Initial Model (Saturated)}
-%\centering
-%\begin{threeparttable}
-%\begin{tabular}[t]{lrrrrr}
-%  \toprule
-%  & Estimate & t value & Pr($>$$|$t$|$) & 95\% CI \\
-%      \midrule
-%    (Intercept) & 0.2844 & 4.17 & $<$0.0001 & (0.1503, 0.4184) \\ 
-%    A & 0.0903 & 1.04 & 0.3004 & (-0.0809, 0.2616) \\ 
-%    GF & 0.0835  & 0.39 & 0.6978 & (-0.3392, 0.5063) \\ 
-%    Plus-Minus & -0.1309 & -2.65 & 0.0085 & (-0.2282, -0.0336) \\ 
-%    TOI & 1.0859 & 2.85 & 0.0046 & (0.3375, 1.8343) \\ 
-%    iSF & 0.1505 & 1.88 & 0.0608 & (-0.0068, 0.3077) \\ 
-%    Yrs\_in\_NHL & 0.2537  & 8.65 & <0.0001 & (0.1960, 0.3114) \\ 
-%    Wt & 0.0927 & 3.17 & 0.0016 & (0.0352, 0.1501) \\ 
-%    GP & -1.2380 & -8.37 & <0.0001 & (-1.5289, -0.9470) \\ 
-%    PDO & 0.2387 & 1.81 & 0.0714 & (-0.0209, 0.4983) \\ 
-%    CA & 1.2701 & 2.16 & 0.0311 & (0.1157, 2.4244) \\ 
-%    CF & 0.6446 & 0.85 & 0.3961 & (-0.8473, 2.1364) \\
-%    FA & -1.0774 & -1.71 & 0.0882 & (-2.3167, 0.1619) \\ 
-%    FF & 0.2472 & 0.29 & 0.7737 & (-1.4420, 1.9363) \\ 
-%    xGA & -0.1335 & -0.55 & 0.5807 & (-0.6084, 0.3414) \\ 
-%    xGF & -0.6520 & -2.12 & 0.0347 & (-1.2568, -0.0472) \\ 
-%    CA:CF & -0.1448 & -0.30 & 0.7653 & (-1.0980, 0.8084) \\ 
-%    FA:FF & 0.6112  & 1.20 & 0.2314 & (-0.3915, 1.6139) \\ 
-%    CA:FA & -0.5539 & -3.39 & 0.0008 & (-0.8752, -0.2326) \\ 
-%    CF:FF & -0.4170 & -2.84 & 0.0047 & (-0.7056, -0.1284) \\ 
-%    xGA:xGF & 0.2437 & 1.33 & 0.1842 & (-0.1165, 0.6038) \\ 
-%    PDO:xGA & 0.0339 & 0.25 & 0.8040 & (-0.2346, 0.3025) \\ 
-%    PDO:xGF & 0.1089 & 0.71 & 0.4789 & (-0.1932, 0.4111) \\ 
-%   \bottomrule
-%\end{tabular}
-%    \begin{tablenotes}
-%      \item $R^2_a$ = 0.72
-%    \end{tablenotes}
-%  \end{threeparttable}    
-%\end{table}
-
-% Remove CI
 \begin{table}[tph]
@@
 
-%\begin{table}
-%\centering
-%\caption{Basis Dimension Check, Initial GAM} 
-%\label{tab.basis.check.initial}
-%\begin{tabular}[t]{lrrrr}
-%\hline
-%& k\textquotesingle & edf & k-index & p-value\\
-%\hline
-%s(G) & 9.00 & 1.00 & 1.00 & 0.490\\
-%s(A) & 9.00 & 1.00 & 1.09 & 0.965\\
-%s(GF) & 9.00 & 1.00 & 0.97 & 0.260\\
-%s(plus\_mins) & 9.00 & 1.15 & 0.93 & 0.045\\
-%s(TOI) & 9.00 & 3.85 & 0.99 &  0.420\\
-%s(iSF) & 9.00 & 1.00 & 1.02 & 0.615\\
-%s(Yrs\_in\_NHL) & 9.00 & 3.29 & 0.98 & 0.310\\
-%s(Wt) & 9.00 & 2.03 & 1.04 & 0.760\\
-%s(GP) & 9.00 & 1.00 & 0.96 & 0.195\\
-%\hline
-%\end{tabular}
-%\end{table}
-
-%\begin{table}
-%\centering
-%\caption{Basis Dimension Check, Revised GAM} 
-%\label{tab.basis.check.revised}
-%\begin{tabular}[t]{lrrrr}
-%\hline
-%& k\textquotesingle & edf & k-index & p-value\\
-%\hline
-%s(TOI) & 9.00 & 3.80 & 0.99 &  0.420\\
-%s(Yrs\_in\_NHL) & 9.00 & 3.01 & 0.97 & 0.280\\
-%\hline
-%\end{tabular}
-%\end{table}
 \vfill\break
